---
title: "Fixed-effects analysis of isotope-incorporation kinetics"
date: "2026-09-09"
output: md_document
---

This report uses the corrected combined d3-palmitate and d7-dhSph incorporation workbook. The stored fractional-incorporation measurements are multiplied by 100 only for presentation as percent incorporation. The underlying incorporation metric is not recalculated.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

input_xlsx <- Sys.getenv(
  "INCORPORATION_XLSX",
  unset = file.path("data", "10. Final Incorporation rate _ Normalized to each class total SL_d3-pal and d7-dhsph_AP_5-09-2026.xlsx")
)
output_dir <- Sys.getenv(
  "KINETIC_OUTPUT_DIR",
  unset = file.path("outputs", "isotope_incorporation_kinetics")
)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
trajectory_dir <- file.path(output_dir, "fitted_trajectories")
diagnostic_dir <- file.path(output_dir, "diagnostic_plots")
dir.create(trajectory_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)

if (!requireNamespace("readxl", quietly = TRUE)) stop("The readxl package is required.")

time_levels <- c(0.5, 1, 2, 4, 8)
blue <- "#0000FF"
red <- "#FF0000"
group_colours <- c(Uninfected = blue, Infected = red)
group_symbols <- c(Uninfected = 16, Infected = 15)
mean_point_cex <- 2.0
axis_line_lwd <- 7.0
axis_tick_lwd <- 3.5
text_scale <- 2.0
axis_text_cex <- 1.0 * text_scale
axis_text_font <- 2
axis_label_cex <- 0.7 * axis_text_cex
title_cex <- 0.7 * text_scale
title_font <- 2
legend_cex <- 0.82 * text_scale
fdr_cex <- 1.0 * text_scale
group_display_labels <- c(Uninfected = "-Sne", Infected = "+Sne")

format_axis_labels <- function(x) {
  labels <- ifelse(
    abs(x - round(x)) < .Machine$double.eps^0.5,
    as.character(as.integer(round(x))),
    sprintf("%.1f", x)
  )
  sub("\\.0$", "", labels)
}

y_axis_overrides <- list(
  d7_SMS_SM = list(limits = c(0, 3.0), breaks = seq(0, 3.0, by = 1)),
  d3_LacCerS_LacCer = list(limits = c(0, 4.0), breaks = seq(0, 4.0, by = 1.0)),
  d7_LacCerS_LacCer = list(limits = c(0, 2.0), breaks = seq(0, 2.0, by = 1.0))
)

get_y_axis_settings <- function(endpoint_id, default_limits) {
  override <- y_axis_overrides[[endpoint_id]]
  if (is.null(override)) {
    y_upper_needed <- max(0, default_limits, na.rm = TRUE)
    if (!is.finite(y_upper_needed) || y_upper_needed <= 0) y_upper_needed <- 1

    breaks <- pretty(c(0, y_upper_needed), n = 5)
    breaks <- breaks[breaks >= 0]
    if (length(breaks) == 0) breaks <- c(0, y_upper_needed)
    if (max(breaks) < y_upper_needed) {
      step <- if (length(breaks) >= 2) {
        tail(breaks, 1) - tail(breaks, 2)[1]
      } else {
        y_upper_needed
      }
      breaks <- c(breaks, max(breaks) + step)
    }
    breaks <- unique(breaks)
    return(list(
      limits = c(0, max(breaks)),
      breaks = breaks,
      labels = format_axis_labels(breaks)
    ))
  }
  list(
    limits = c(override$limits[1], tail(override$breaks, 1)),
    breaks = override$breaks,
    labels = format_axis_labels(override$breaks)
  )
}

draw_prism_axes <- function(x_at, x_labels, y_at = NULL, y_labels = NULL) {
  usr <- par("usr")
  segments(usr[1], usr[3], usr[2], usr[3], col = "black", lwd = axis_line_lwd)
  segments(usr[1], usr[3], usr[1], usr[4], col = "black", lwd = axis_line_lwd)
  axis(1, at = x_at, labels = x_labels, col = "black", col.ticks = "black",
       lwd = 0, lwd.ticks = axis_tick_lwd, cex.axis = axis_text_cex,
       font = axis_text_font, gap.axis = -1)
  if (is.null(y_at)) {
    axis(2, las = 1, col = "black", col.ticks = "black",
         lwd = 0, lwd.ticks = axis_tick_lwd, cex.axis = axis_text_cex,
         font = axis_text_font)
  } else {
    axis(2, at = y_at, labels = y_labels, las = 1,
         col = "black", col.ticks = "black",
         lwd = 0, lwd.ticks = axis_tick_lwd, cex.axis = axis_text_cex,
         font = axis_text_font)
  }
}

draw_t_error_bars <- function(x, ymin, ymax, col, width = 0.06, lwd = 2.5) {
  arrows(
    x0 = x, y0 = ymin, x1 = x, y1 = ymax,
    angle = 90, code = 3, length = width,
    col = col, lwd = lwd
  )
}

draw_group_legend <- function() {
  usr <- par("usr")
  old_xpd <- par("xpd")
  on.exit(par(xpd = old_xpd), add = TRUE)
  par(xpd = NA)
  legend(
    x = usr[2] + 0.005 * diff(usr[1:2]),
    y = usr[4],
    legend = unname(group_display_labels[c("Uninfected", "Infected")]),
    col = c(blue, red),
    lwd = 2.5,
    pch = c(group_symbols[["Uninfected"]], group_symbols[["Infected"]]),
    pt.cex = mean_point_cex,
    seg.len = 1.2,
    x.intersp = 0.8,
    bty = "n",
    cex = legend_cex,
    xjust = 0,
    yjust = 1
  )
}

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 0.001) return("<0.001")
  sprintf("%.3f", p)
}

fmt_fdr <- function(p) {
  if (is.na(p)) return("FDR = NA")
  if (p < 0.001) return("FDR < 0.001")
  paste0("FDR = ", sprintf("%.3f", p))
}

safe_name <- function(x) {
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  gsub("^_|_$", "", x)
}

meta_d3 <- data.frame(
  tracer = "d3-palmitate",
  sheet = "d3- Palmitate_ IR",
  source_endpoint = c(
    "Labelled d3-dhsph Incorporation rate (SPT)",
    "Labelled (d3+d6) dhCer Incorporation rate (CerS)",
    "Labelled (d3+d6) Cer Incorporation rate (DES)",
    "Labelled (d3+d6) dhSM Incorporation rate (SMS)",
    "Labelled (d3+d6) SM Incorporation rate (SMS)",
    "Labelled d3-sph Incorporation rate (AC)",
    "Labelled (d3+d6) HexCer Incorporation rate (GCS)",
    "Labelled (d3+d6) LacCer Incorporation rate (LacCerS)"
  ),
  endpoint_id = c(
    "d3_SPT_dhSph", "d3_CerS_dhCer", "d3_DES_Cer", "d3_SMS_dhSM",
    "d3_SMS_SM", "d3_AC_Sph", "d3_GCS_HexCer", "d3_LacCerS_LacCer"
  ),
  display_endpoint = c(
    "SPT-linked d3-dhSph", "CerS-linked labelled dhCer", "DES-linked labelled Cer",
    "SMS-linked labelled dhSM", "SMS-linked labelled SM", "AC-linked d3-Sph",
    "GCS-linked labelled HexCer", "LacCerS-linked labelled LacCer"
  ),
  primary_endpoint = TRUE,
  stringsAsFactors = FALSE
)

meta_d7 <- data.frame(
  tracer = "d7-dhSph",
  sheet = "d7-dhSph_IR",
  source_endpoint = c(
    "Labelled d7-dhsph Incorporation rate (SPT)",
    "Labelled d7 dhCer Incorporation rate (CerS)",
    "Labelled d7 Cer Incorporation rate (DES)",
    "Labelled d7 dhSM Incorporation rate (SMS)",
    "Labelled d7 SM Incorporation rate (SMS)",
    "Labelled d7 HexCer Incorporation rate (GCS)",
    "Labelled d7 LacCer Incorporation rate (LacCerS)",
    "Labelled d7-sph Incorporation rate (AC)",
    "Labelled d7-S1P Incorporation rate (SpK)"
  ),
  endpoint_id = c(
    "d7_precursor_dhSph", "d7_CerS_dhCer", "d7_DES_Cer", "d7_SMS_dhSM",
    "d7_SMS_SM", "d7_GCS_HexCer", "d7_LacCerS_LacCer", "d7_AC_Sph", "d7_SpK_S1P"
  ),
  display_endpoint = c(
    "d7-dhSph precursor-pool labelling", "CerS-linked d7-dhCer", "DES-linked d7-Cer",
    "SMS-linked d7-dhSM", "SMS-linked d7-SM", "GCS-linked d7-HexCer",
    "LacCerS-linked d7-LacCer", "AC-linked d7-Sph", "SpK-linked d7-S1P"
  ),
  primary_endpoint = c(FALSE, rep(TRUE, 8)),
  stringsAsFactors = FALSE
)

endpoint_meta <- rbind(meta_d3, meta_d7)
endpoint_meta$plot_index <- seq_len(nrow(endpoint_meta))
endpoint_meta$plot_file <- sprintf(
  "%02d_%s.pdf", endpoint_meta$plot_index,
  vapply(endpoint_meta$endpoint_id, safe_name, character(1))
)

plot_endpoint_meta <- endpoint_meta[endpoint_meta$endpoint_id != "d7_precursor_dhSph", , drop = FALSE]
plot_endpoint_meta$plot_index <- seq_len(nrow(plot_endpoint_meta))
plot_endpoint_meta$plot_file <- sprintf(
  "%02d_%s.pdf", plot_endpoint_meta$plot_index,
  vapply(plot_endpoint_meta$endpoint_id, safe_name, character(1))
)

extract_match <- function(x, pattern) {
  hit <- regexpr(pattern, x, perl = TRUE)
  out <- rep(NA_character_, length(x))
  ok <- hit > 0
  out[ok] <- regmatches(x, hit)[ok]
  out
}

read_tracer_sheet <- function(meta) {
  sheet_name <- unique(meta$sheet)
  tracer_name <- unique(meta$tracer)
  tracer_pattern <- if (tracer_name == "d3-palmitate") "d3-Palmitate" else "d7-dhSph"
  x <- readxl::read_excel(input_xlsx, sheet = sheet_name)

  if (!identical(names(x)[1], "Würzburg sample ID")) {
    stop(sprintf("The first column of '%s' must be 'Würzburg sample ID'.", sheet_name))
  }
  missing_endpoints <- setdiff(meta$source_endpoint, names(x))
  if (length(missing_endpoints)) {
    stop(sprintf("Missing expected columns in '%s': %s", sheet_name, paste(missing_endpoints, collapse = "; ")))
  }

  sample_ids_all <- as.character(x[["Würzburg sample ID"]])
  tracer_rows <- grepl(tracer_pattern, sample_ids_all, fixed = TRUE)
  control_rows <- !tracer_rows
  endpoint_all <- as.data.frame(lapply(x[meta$source_endpoint], as.numeric), check.names = FALSE)

  qc <- data.frame(
    tracer = tracer_name,
    sheet = sheet_name,
    total_rows = nrow(x),
    tracer_rows = sum(tracer_rows),
    no_label_control_rows = sum(control_rows),
    duplicate_sample_ids = sum(duplicated(sample_ids_all)),
    missing_or_nonfinite_endpoint_cells = sum(!is.finite(as.matrix(endpoint_all))),
    values_outside_zero_to_one = sum(as.matrix(endpoint_all) < 0 | as.matrix(endpoint_all) > 1, na.rm = TRUE),
    nonzero_control_cells = sum(as.matrix(endpoint_all[control_rows, , drop = FALSE]) != 0, na.rm = TRUE),
    stringsAsFactors = FALSE
  )

  x <- x[tracer_rows, , drop = FALSE]
  sample_ids <- sample_ids_all[tracer_rows]
  time_text <- extract_match(sample_ids, "(0\\.5|1|2|4|8)h")
  replicate_text <- extract_match(sample_ids, "R[123]")
  design <- data.frame(
    sample_name = sample_ids,
    infection = factor(
      ifelse(grepl("Sne", sample_ids, fixed = TRUE), "Infected", "Uninfected"),
      levels = c("Uninfected", "Infected")
    ),
    time_h = as.numeric(sub("h", "", time_text, fixed = TRUE)),
    replicate = factor(replicate_text, levels = c("R1", "R2", "R3")),
    stringsAsFactors = FALSE
  )

  if (nrow(design) != 30L) stop(sprintf("Expected 30 tracer rows in '%s'; found %d.", sheet_name, nrow(design)))
  if (anyNA(design$time_h) || anyNA(design$replicate)) stop(sprintf("Could not parse every time/replicate in '%s'.", sheet_name))
  design_counts <- table(design$infection, design$time_h)
  if (!all(dim(design_counts) == c(2L, 5L)) || any(design_counts != 3L)) {
    stop(sprintf("Incomplete infection-by-time design in '%s'.", sheet_name))
  }

  long_rows <- lapply(seq_len(nrow(meta)), function(i) {
    m <- meta[i, , drop = FALSE]
    data.frame(
      tracer = tracer_name,
      endpoint_id = m$endpoint_id,
      display_endpoint = m$display_endpoint,
      source_endpoint = m$source_endpoint,
      primary_endpoint = m$primary_endpoint,
      sample_name = design$sample_name,
      infection = design$infection,
      time_h = design$time_h,
      replicate = design$replicate,
      incorporation_fraction = as.numeric(x[[m$source_endpoint]]),
      stringsAsFactors = FALSE
    )
  })
  out <- do.call(rbind, long_rows)
  out$infection <- factor(out$infection, levels = c("Uninfected", "Infected"))
  out$replicate <- factor(out$replicate, levels = c("R1", "R2", "R3"))
  out$incorporation_percent <- 100 * out$incorporation_fraction
  out$time_c <- out$time_h - 0.5
  out$log2_time_c <- log2(out$time_h) - log2(0.5)
  out$time_factor <- factor(out$time_h, levels = time_levels)
  list(data = out, qc = qc)
}

d3_import <- read_tracer_sheet(meta_d3)
d7_import <- read_tracer_sheet(meta_d7)
analysis_data <- rbind(d3_import$data, d7_import$data)
input_qc <- rbind(d3_import$qc, d7_import$qc)

if (any(!is.finite(analysis_data$incorporation_fraction))) stop("Non-finite incorporation value found.")
if (any(analysis_data$incorporation_fraction < 0 | analysis_data$incorporation_fraction > 1)) stop("Incorporation fraction outside 0-1.")

write.csv(analysis_data, file.path(output_dir, "incorporation_observations_used.csv"), row.names = FALSE)
write.csv(input_qc, file.path(output_dir, "input_quality_control.csv"), row.names = FALSE)

group_summary_raw <- aggregate(
  incorporation_percent ~ tracer + endpoint_id + display_endpoint + primary_endpoint + infection + time_h,
  data = analysis_data,
  FUN = function(x) c(mean = mean(x), sd = sd(x), n = length(x))
)
group_summary <- data.frame(
  tracer = group_summary_raw$tracer,
  endpoint_id = group_summary_raw$endpoint_id,
  display_endpoint = group_summary_raw$display_endpoint,
  primary_endpoint = group_summary_raw$primary_endpoint,
  infection = group_summary_raw$infection,
  time_h = group_summary_raw$time_h,
  mean_percent = group_summary_raw$incorporation_percent[, "mean"],
  sd_percent = group_summary_raw$incorporation_percent[, "sd"],
  n = group_summary_raw$incorporation_percent[, "n"],
  stringsAsFactors = FALSE
)
group_summary <- group_summary[order(group_summary$tracer, group_summary$endpoint_id, group_summary$infection, group_summary$time_h), ]
write.csv(group_summary, file.path(output_dir, "incorporation_group_summary.csv"), row.names = FALSE)

fixed_formula <- function(model_type, full = TRUE) {
  if (model_type == "linear_time") {
    if (full) incorporation_percent ~ infection * time_c else incorporation_percent ~ infection + time_c
  } else if (model_type == "log2_time") {
    if (full) incorporation_percent ~ infection * log2_time_c else incorporation_percent ~ infection + log2_time_c
  } else if (model_type == "natural_spline_df2") {
    if (full) incorporation_percent ~ infection * splines::ns(time_h, df = 2) else incorporation_percent ~ infection + splines::ns(time_h, df = 2)
  } else if (model_type == "categorical_time") {
    if (full) incorporation_percent ~ infection * time_factor else incorporation_percent ~ infection + time_factor
  } else {
    stop("Unknown model type: ", model_type)
  }
}

fit_lm <- function(d, model_type, full = TRUE) {
  lm(fixed_formula(model_type, full = full), data = d)
}

lack_of_fit_p <- function(model, categorical_model) {
  as.numeric(anova(model, categorical_model)$"Pr(>F)"[2])
}

interaction_p_lm <- function(d, model_type) {
  reduced <- fit_lm(d, model_type, full = FALSE)
  full <- fit_lm(d, model_type, full = TRUE)
  as.numeric(anova(reduced, full)$"Pr(>F)"[2])
}

select_kinetic_model <- function(d) {
  d <- droplevels(d)
  model_types <- c("linear_time", "log2_time", "natural_spline_df2", "categorical_time")
  models <- setNames(lapply(model_types, function(z) fit_lm(d, z, full = TRUE)), model_types)
  lof <- c(
    linear_time = lack_of_fit_p(models$linear_time, models$categorical_time),
    log2_time = lack_of_fit_p(models$log2_time, models$categorical_time),
    natural_spline_df2 = lack_of_fit_p(models$natural_spline_df2, models$categorical_time),
    categorical_time = NA_real_
  )
  adequate <- is.na(lof) | lof >= 0.05
  aic_linear <- AIC(models$linear_time)
  aic_log2 <- AIC(models$log2_time)

  if (adequate["linear_time"] && !adequate["log2_time"]) {
    final_type <- "linear_time"
    justification <- "Clock time was adequate whereas log2(time) was not."
  } else if (!adequate["linear_time"] && adequate["log2_time"]) {
    final_type <- "log2_time"
    justification <- "log2(time) was adequate whereas clock time was not."
  } else if (adequate["linear_time"] && adequate["log2_time"] && aic_log2 <= aic_linear - 2) {
    final_type <- "log2_time"
    justification <- "Both simple time scales were adequate and log2(time) had at least 2 AIC units of support."
  } else if (adequate["linear_time"] || adequate["log2_time"]) {
    final_type <- "linear_time"
    justification <- "The simple models were adequate without clear AIC support for log2(time); clock time was retained for parsimony."
  } else if (adequate["natural_spline_df2"]) {
    final_type <- "natural_spline_df2"
    justification <- "Both simple continuous-time models lacked fit; a two-degree-of-freedom natural spline removed detectable lack of fit."
  } else {
    final_type <- "categorical_time"
    justification <- "Linear time, log2(time), and the low-complexity spline all lacked fit; unconstrained categorical time was retained."
  }

  list(
    models = models,
    lack_of_fit = lof,
    adequate = adequate,
    final_type = final_type,
    final_model = models[[final_type]],
    justification = justification
  )
}

contrast_lm <- function(model, L, level = 0.95) {
  b <- coef(model)
  V <- vcov(model)
  if (is.null(names(L))) names(L) <- names(b)
  L <- L[names(b)]
  estimate <- as.numeric(sum(L * b))
  se <- as.numeric(sqrt(t(L) %*% V %*% L))
  df <- df.residual(model)
  statistic <- estimate / se
  p_value <- 2 * pt(-abs(statistic), df = df)
  critical <- qt(1 - (1 - level) / 2, df = df)
  data.frame(
    estimate = estimate, std_error = se, df = df, statistic = statistic,
    p_value = p_value, conf_low = estimate - critical * se,
    conf_high = estimate + critical * se
  )
}

fixed_model_matrix <- function(model_type, newdata) {
  model.matrix(delete.response(terms(fixed_formula(model_type, full = TRUE))), newdata)
}

newdata_contrast <- function(model, model_type, newdata_a, newdata_b) {
  Xa <- fixed_model_matrix(model_type, newdata_a)
  Xb <- fixed_model_matrix(model_type, newdata_b)
  bnames <- names(coef(model))
  L <- as.numeric(Xa[1, bnames, drop = TRUE] - Xb[1, bnames, drop = TRUE])
  names(L) <- bnames
  contrast_lm(model, L)
}

extract_slopes <- function(model, model_type) {
  if (!model_type %in% c("linear_time", "log2_time")) return(NULL)
  term_time <- if (model_type == "linear_time") "time_c" else "log2_time_c"
  term_interaction <- paste0("infectionInfected:", term_time)
  bnames <- names(coef(model))
  L_u <- setNames(rep(0, length(bnames)), bnames)
  L_i <- L_u
  L_d <- L_u
  L_u[term_time] <- 1
  L_i[term_time] <- 1
  L_i[term_interaction] <- 1
  L_d[term_interaction] <- 1
  list(
    uninfected = contrast_lm(model, L_u),
    infected = contrast_lm(model, L_i),
    difference = contrast_lm(model, L_d)
  )
}

model_fit_stats <- function(model) {
  s <- summary(model)
  c(R2 = unname(s$r.squared), adjusted_R2 = unname(s$adj.r.squared), residual_sd = unname(s$sigma))
}

prediction_frame <- function(model, model_type, times = NULL) {
  if (is.null(times)) {
    times <- if (model_type == "categorical_time") time_levels else seq(0.5, 8, length.out = 200)
  }

  nd <- expand.grid(
    time_h = times,
    infection = factor(c("Uninfected", "Infected"), levels = c("Uninfected", "Infected"))
  )
  nd$time_c <- nd$time_h - 0.5
  nd$log2_time_c <- log2(nd$time_h) - log2(0.5)
  nd$time_factor <- factor(nd$time_h, levels = time_levels)
  X <- fixed_model_matrix(model_type, nd)
  b <- coef(model)
  X <- X[, names(b), drop = FALSE]
  V <- vcov(model)
  nd$fit <- as.numeric(X %*% b)
  nd$se <- sqrt(pmax(0, diag(X %*% V %*% t(X))))
  critical <- qt(0.975, df = df.residual(model))
  nd$lwr <- nd$fit - critical * nd$se
  nd$upr <- nd$fit + critical * nd$se
  nd
}

model_label <- function(model_type) {
  switch(
    model_type,
    linear_time = "Linear clock time",
    log2_time = "Linear log2(time)",
    natural_spline_df2 = "Natural spline (2 df)",
    categorical_time = "Unconstrained categorical time"
  )
}

fits <- list()
comparison_rows <- list()
coefficient_rows <- list()
summary_rows <- list()
diagnostic_rows <- list()
prediction_rows <- list()

for (endpoint_id in endpoint_meta$endpoint_id) {
  d <- droplevels(analysis_data[analysis_data$endpoint_id == endpoint_id, ])
  m <- endpoint_meta[endpoint_meta$endpoint_id == endpoint_id, , drop = FALSE]
  fit_result <- select_kinetic_model(d)
  fits[[endpoint_id]] <- fit_result

  for (candidate in names(fit_result$models)) {
    mod <- fit_result$models[[candidate]]
    fit_stats <- model_fit_stats(mod)
    comparison_rows[[length(comparison_rows) + 1L]] <- data.frame(
      tracer = m$tracer,
      endpoint_id = endpoint_id,
      display_endpoint = m$display_endpoint,
      candidate_model = candidate,
      selected = candidate == fit_result$final_type,
      AIC = AIC(mod),
      BIC = BIC(mod),
      log_likelihood = as.numeric(logLik(mod)),
      lack_of_fit_vs_categorical_p = fit_result$lack_of_fit[candidate],
      lack_of_fit_acceptable = fit_result$adequate[candidate],
      R2 = fit_stats["R2"],
      adjusted_R2 = fit_stats["adjusted_R2"],
      stringsAsFactors = FALSE
    )
  }

  final_model <- fit_result$final_model
  interaction_p <- interaction_p_lm(d, fit_result$final_type)
  fit_stats <- model_fit_stats(final_model)
  standardized_residuals <- as.numeric(rstandard(final_model))
  fitted_values <- as.numeric(fitted(final_model))
  shapiro_p <- if (length(unique(standardized_residuals)) >= 3L) shapiro.test(standardized_residuals)$p.value else NA_real_
  bp_aux <- lm(I(standardized_residuals^2) ~ fitted_values)
  bp_stat <- nrow(d) * summary(bp_aux)$r.squared
  bp_p <- pchisq(bp_stat, df = 1, lower.tail = FALSE)
  cook <- cooks.distance(final_model)
  cook_threshold <- 4 / nrow(d)
  influential <- which(cook > cook_threshold)

  baseline_u <- data.frame(
    infection = factor("Uninfected", levels = c("Uninfected", "Infected")),
    time_h = 0.5, time_c = 0, log2_time_c = 0,
    time_factor = factor(0.5, levels = time_levels)
  )
  baseline_i <- baseline_u
  baseline_i$infection <- factor("Infected", levels = c("Uninfected", "Infected"))
  baseline_effect <- newdata_contrast(final_model, fit_result$final_type, baseline_i, baseline_u)
  slopes <- extract_slopes(final_model, fit_result$final_type)

  summary_rows[[length(summary_rows) + 1L]] <- data.frame(
    tracer = m$tracer,
    endpoint_id = endpoint_id,
    display_endpoint = m$display_endpoint,
    source_endpoint = m$source_endpoint,
    primary_endpoint = m$primary_endpoint,
    selected_model = fit_result$final_type,
    selected_model_label = model_label(fit_result$final_type),
    infection_by_time_interaction_p = interaction_p,
    infection_effect_at_0_5h = baseline_effect$estimate,
    infection_effect_at_0_5h_conf_low = baseline_effect$conf_low,
    infection_effect_at_0_5h_conf_high = baseline_effect$conf_high,
    infection_effect_at_0_5h_p = baseline_effect$p_value,
    R2 = fit_stats["R2"],
    adjusted_R2 = fit_stats["adjusted_R2"],
    residual_sd = fit_stats["residual_sd"],
    selection_justification = fit_result$justification,
    slope_interpretation_valid = fit_result$final_type %in% c("linear_time", "log2_time"),
    slope_uninfected = if (is.null(slopes)) NA_real_ else slopes$uninfected$estimate,
    slope_infected = if (is.null(slopes)) NA_real_ else slopes$infected$estimate,
    slope_difference = if (is.null(slopes)) NA_real_ else slopes$difference$estimate,
    slope_difference_conf_low = if (is.null(slopes)) NA_real_ else slopes$difference$conf_low,
    slope_difference_conf_high = if (is.null(slopes)) NA_real_ else slopes$difference$conf_high,
    slope_difference_p = if (is.null(slopes)) NA_real_ else slopes$difference$p_value,
    slope_unit = if (fit_result$final_type == "linear_time") "percentage points per hour" else if (fit_result$final_type == "log2_time") "percentage points per doubling of time" else NA_character_,
    stringsAsFactors = FALSE
  )

  tt <- summary(final_model)$coefficients
  ci <- confint(final_model)
  coefficient_rows[[length(coefficient_rows) + 1L]] <- data.frame(
    tracer = m$tracer,
    endpoint_id = endpoint_id,
    display_endpoint = m$display_endpoint,
    selected_model = fit_result$final_type,
    term = rownames(tt),
    estimate = tt[, "Estimate"],
    std_error = tt[, "Std. Error"],
    df = df.residual(final_model),
    statistic = tt[, "t value"],
    p_value = tt[, "Pr(>|t|)"],
    conf_low = ci[, 1],
    conf_high = ci[, 2],
    row.names = NULL,
    stringsAsFactors = FALSE
  )

  diagnostic_rows[[length(diagnostic_rows) + 1L]] <- data.frame(
    tracer = m$tracer,
    endpoint_id = endpoint_id,
    display_endpoint = m$display_endpoint,
    selected_model = fit_result$final_type,
    n_observations = nrow(d),
    exact_zero_count = sum(d$incorporation_percent == 0),
    Shapiro_Wilk_p = shapiro_p,
    residual_normality_assessment = ifelse(!is.na(shapiro_p) && shapiro_p < 0.05, "Potential deviation", "Acceptable"),
    Breusch_Pagan_like_p = bp_p,
    variance_assessment = ifelse(!is.na(bp_p) && bp_p < 0.05, "Potential heteroscedasticity", "Acceptable"),
    standardized_residual_count_abs_gt_2_5 = sum(abs(standardized_residuals) > 2.5),
    approximate_Cooks_threshold = cook_threshold,
    approximate_influential_observation_count = length(influential),
    approximate_influential_sample_names = if (length(influential)) paste(d$sample_name[influential], collapse = "; ") else "None",
    R2 = fit_stats["R2"],
    adjusted_R2 = fit_stats["adjusted_R2"],
    residual_sd = fit_stats["residual_sd"],
    stringsAsFactors = FALSE
  )

  pred <- prediction_frame(final_model, fit_result$final_type)
  pred$tracer <- m$tracer
  pred$endpoint_id <- endpoint_id
  pred$display_endpoint <- m$display_endpoint
  pred$selected_model <- fit_result$final_type
  prediction_rows[[length(prediction_rows) + 1L]] <- pred[, c(
    "tracer", "endpoint_id", "display_endpoint", "selected_model",
    "infection", "time_h", "fit", "se", "lwr", "upr"
  )]
}

model_comparison <- do.call(rbind, comparison_rows)
model_coefficients <- do.call(rbind, coefficient_rows)
model_summary <- do.call(rbind, summary_rows)
model_diagnostics <- do.call(rbind, diagnostic_rows)
fitted_predictions <- do.call(rbind, prediction_rows)

model_summary$interaction_p_adjusted_BH <- NA_real_
for (tr in unique(model_summary$tracer)) {
  idx <- model_summary$tracer == tr & model_summary$primary_endpoint
  model_summary$interaction_p_adjusted_BH[idx] <- p.adjust(
    model_summary$infection_by_time_interaction_p[idx], method = "BH"
  )
}
model_summary$primary_conclusion <- ifelse(
  !model_summary$primary_endpoint,
  "Precursor-pool quality-control endpoint; excluded from the primary enzyme-linked multiplicity family.",
  ifelse(
    model_summary$interaction_p_adjusted_BH < 0.05,
    "Evidence of an infection-dependent incorporation trajectory after BH correction.",
    "No infection-dependent incorporation trajectory detected after BH correction."
  )
)

write.csv(model_summary, file.path(output_dir, "kinetic_model_summary.csv"), row.names = FALSE)
write.csv(model_comparison, file.path(output_dir, "kinetic_model_comparison.csv"), row.names = FALSE)
write.csv(model_coefficients, file.path(output_dir, "kinetic_model_coefficients.csv"), row.names = FALSE)
write.csv(model_diagnostics, file.path(output_dir, "kinetic_model_diagnostics.csv"), row.names = FALSE)
write.csv(fitted_predictions, file.path(output_dir, "kinetic_model_fitted_predictions.csv"), row.names = FALSE)

plot_trajectory <- function(d, fit_result, meta_row) {
  model <- fit_result$final_model
  model_type <- fit_result$final_type
  pred <- prediction_frame(model, model_type)
  pred_timepoints <- prediction_frame(model, model_type, times = time_levels)
  point_summary <- group_summary[group_summary$endpoint_id == meta_row$endpoint_id, ]
  summary_row <- model_summary[model_summary$endpoint_id == meta_row$endpoint_id, ]
  y_min <- min(0, d$incorporation_percent, pred$lwr, na.rm = TRUE)
  y_max <- max(d$incorporation_percent, pred$upr, na.rm = TRUE)
  y_pad <- max(0.2, 0.18 * (y_max - y_min))
  y_axis <- get_y_axis_settings(as.character(meta_row$endpoint_id), c(y_min, y_max + y_pad))

  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)
  par(mar = c(8.0, 8.7, 6.3, 11.0), mgp = c(5.5, 1.6, 0))
  plot(
    NA, xlim = c(0.35, 8.15), ylim = y_axis$limits,
    xlab = "Time after label addition (h)", ylab = "Fractional incorporation (%)",
    main = meta_row$display_endpoint, axes = FALSE, frame.plot = FALSE,
    xaxs = "i", yaxs = "i",
    cex.lab = axis_label_cex, cex.main = title_cex, font.main = title_font
  )
  draw_prism_axes(
    time_levels, c("0.5", "1", "2", "4", "8"),
    y_at = y_axis$breaks, y_labels = y_axis$labels
  )

  for (grp in c("Uninfected", "Infected")) {
    p <- pred[pred$infection == grp, ]
    p_timepoints <- pred_timepoints[pred_timepoints$infection == grp, ]
    col <- group_colours[[grp]]

    if (model_type == "natural_spline_df2") {
      polygon(
        c(p$time_h, rev(p$time_h)), c(p$lwr, rev(p$upr)),
        col = grDevices::adjustcolor(col, alpha.f = 0.17), border = NA
      )
      lines(p$time_h, p$fit, col = col, lwd = 2.5)
      next
    }

    lines(p$time_h, p$fit, col = col, lwd = 2.5)
    draw_t_error_bars(p_timepoints$time_h, p_timepoints$lwr, p_timepoints$upr, col = col)
  }

  for (grp in c("Uninfected", "Infected")) {
    p <- point_summary[as.character(point_summary$infection) == grp, ]
    points(
      p$time_h, p$mean_percent,
      pch = group_symbols[[grp]], col = group_colours[[grp]], cex = mean_point_cex
    )
  }
  draw_group_legend()

  note <- if (summary_row$primary_endpoint) fmt_fdr(summary_row$interaction_p_adjusted_BH) else "FDR = NA"
  usr <- par("usr")
  text(
    usr[2] - 0.015 * diff(usr[1:2]),
    usr[4] - 0.020 * diff(usr[3:4]),
    labels = note, adj = c(1, 1), cex = fdr_cex
  )
}

plot_diagnostics <- function(d, fit_result, meta_row) {
  model <- fit_result$final_model
  r <- as.numeric(rstandard(model))
  f <- as.numeric(fitted(model))
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)
  par(
    mfrow = c(2, 2), mar = c(5.0, 5.0, 4.0, 1.0), oma = c(0, 0, 4.0, 0),
    cex.axis = axis_text_cex, cex.lab = axis_label_cex, cex.main = title_cex,
    font.axis = axis_text_font, font.main = title_font
  )
  plot(f, r, pch = 21, bg = ifelse(d$infection == "Infected", red, blue), col = "black", xlab = "Fitted value", ylab = "Standardized residual", main = "Residuals vs fitted")
  abline(h = 0, lty = 2, col = "grey40")
  qqnorm(r, pch = 21, bg = ifelse(d$infection == "Infected", red, blue), col = "black", main = "Normal Q-Q")
  qqline(r, col = "grey40", lwd = 1.5)
  hist(r, breaks = "FD", col = "grey85", border = "black", main = "Residual distribution", xlab = "Standardized residual")
  plot(d$time_h, r, pch = ifelse(d$infection == "Uninfected", 21, 24), bg = ifelse(d$infection == "Infected", red, blue), col = "black", xlab = "Time after label addition (h)", ylab = "Standardized residual", main = "Residuals over time", xaxt = "n")
  axis(1, at = time_levels, cex.axis = axis_text_cex, font = axis_text_font)
  abline(h = 0, lty = 2, col = "grey40")
  mtext(meta_row$display_endpoint, side = 3, outer = TRUE, line = 0.6, font = title_font, cex = title_cex)
}

for (plot_dir in c(trajectory_dir, diagnostic_dir)) {
  unlink(list.files(plot_dir, pattern = "\\.(png|pdf)$", full.names = TRUE))
}

for (i in seq_len(nrow(plot_endpoint_meta))) {
  m <- plot_endpoint_meta[i, , drop = FALSE]
  d <- droplevels(analysis_data[analysis_data$endpoint_id == m$endpoint_id, ])
  pdf(file.path(trajectory_dir, m$plot_file), width = 11, height = 7, onefile = FALSE)
  plot_trajectory(d, fits[[m$endpoint_id]], m)
  dev.off()
  diagnostic_file <- sub("\\.pdf$", "_diagnostics.pdf", m$plot_file)
  pdf(file.path(diagnostic_dir, diagnostic_file), width = 11, height = 8.5, onefile = FALSE)
  plot_diagnostics(d, fits[[m$endpoint_id]], m)
  dev.off()
}

pdf(file.path(output_dir, "kinetic_model_fitted_trajectories_all_endpoints.pdf"), width = 11, height = 7, onefile = TRUE)
for (i in seq_len(nrow(plot_endpoint_meta))) {
  m <- plot_endpoint_meta[i, , drop = FALSE]
  d <- droplevels(analysis_data[analysis_data$endpoint_id == m$endpoint_id, ])
  plot_trajectory(d, fits[[m$endpoint_id]], m)
}
dev.off()

pdf(file.path(output_dir, "kinetic_model_diagnostic_plots_all_endpoints.pdf"), width = 11, height = 8.5, onefile = TRUE)
for (i in seq_len(nrow(plot_endpoint_meta))) {
  m <- plot_endpoint_meta[i, , drop = FALSE]
  d <- droplevels(analysis_data[analysis_data$endpoint_id == m$endpoint_id, ])
  plot_diagnostics(d, fits[[m$endpoint_id]], m)
}
dev.off()

analysis_log <- c(
  paste("Input workbook:", input_xlsx),
  paste("Input MD5:", unname(tools::md5sum(input_xlsx))),
  paste("Analysis timestamp:", format(Sys.time(), tz = "Europe/Berlin")),
  "Model structure: Gaussian fixed-effects regression with infection, time, and their interaction.",
  "Replicate labels were retained only as observation identifiers; no grouping or repeated-measures structure was fitted.",
  capture.output(sessionInfo())
)
writeLines(analysis_log, file.path(output_dir, "analysis_session_log.txt"))
```

## Design and model rationale

The analysis includes independent cell samples at 0.5, 1, 2, 4, and 8 h, with three biological replicates per infection condition and time point. The replicate labels identify observations but are not used as a grouping variable. All models are ordinary Gaussian fixed-effects regressions, with no repeated-measures temporal correlation.

The response is fractional isotope incorporation expressed as a percentage. Endpoint names use “linked” because these measurements are kinetic pathway proxies rather than direct enzyme-activity assays.

The exogenous d7-dhSph precursor-pool endpoint is retained in statistical summaries as quality control, but excluded from the primary enzyme-linked multiplicity family and from exported plot panels because d7-dhSph bypasses SPT and precursor labelling is close to saturation.

## Input quality control

```{r input-qc, results='asis'}
knitr::kable(input_qc, digits = 3, row.names = FALSE)
```

No-label controls confirm absence of isotope signal and are not included in the kinetic models.

## Model selection

For every endpoint, four Gaussian fixed-effects models were evaluated:

1. `incorporation ~ infection * time`
2. `incorporation ~ infection * log2(time)`
3. `incorporation ~ infection * natural_spline(time, df = 2)`
4. `incorporation ~ infection * categorical_time`

Categorical time served as the unconstrained lack-of-fit reference. If both simple continuous-time models lacked fit, the two-degree-of-freedom natural spline was evaluated. Categorical time was retained when the spline also lacked fit. Selection considered formal lack-of-fit tests, AIC, adjusted R², residual diagnostics, and parsimony.

The global infection-by-time interaction tests whether the incorporation trajectory differs between infected and uninfected samples. A single slope is reported only when a linear clock-time or log2-time model is adequate.

## Primary model results

```{r model-summary, results='asis'}
display_summary <- model_summary[, c(
  "tracer", "display_endpoint", "selected_model_label",
  "infection_by_time_interaction_p", "interaction_p_adjusted_BH",
  "R2", "adjusted_R2", "primary_conclusion"
)]
names(display_summary) <- c(
  "Tracer", "Endpoint", "Selected model", "Interaction P", "BH-adjusted P",
  "R²", "Adjusted R²", "Conclusion"
)
knitr::kable(display_summary, digits = 4, row.names = FALSE)
```

Benjamini-Hochberg adjustment is performed separately within the eight prespecified pathway-linked endpoints for each tracer. The d7-dhSph precursor-pool quality-control endpoint is not assigned a BH-adjusted P value.

## Diagnostics

```{r diagnostics-summary, results='asis'}
diagnostic_display <- model_diagnostics[, c(
  "tracer", "display_endpoint", "selected_model", "exact_zero_count",
  "Shapiro_Wilk_p", "residual_normality_assessment", "Breusch_Pagan_like_p",
  "variance_assessment", "standardized_residual_count_abs_gt_2_5", "adjusted_R2"
)]
names(diagnostic_display) <- c(
  "Tracer", "Endpoint", "Selected model", "Exact zeros", "Shapiro-Wilk P",
  "Normality", "Variance-test P", "Variance", "|Residual| > 2.5", "Adjusted R²"
)
knitr::kable(diagnostic_display, digits = 4, row.names = FALSE)
```

Zeros are retained as biologically plausible observations. Diagnostic flags identify endpoints that require caution but do not automatically justify deleting observations or changing the response distribution.

## Fitted trajectories

Group mean observations, fitted trajectories, 95% confidence intervals, selected model, global interaction P value, BH-adjusted P value, and adjusted R² are shown below.

```{r trajectory-panels, results='asis'}
for (i in seq_len(nrow(plot_endpoint_meta))) {
  cat("\n### ", plot_endpoint_meta$display_endpoint[i], "\n\n", sep = "")
  cat("![](", file.path("fitted_trajectories", plot_endpoint_meta$plot_file[i]), ")\n\n", sep = "")
}
```

## Relation to nonlinear saturation models

A linear model in log2(time) is not itself a logistic or Michaelis-Menten fit. It implies `incorporation = intercept + slope * log2(time)`, which rises rapidly at early clock times and progressively flattens but has no finite plateau. A logistic curve fitted against log2(time) is mathematically related to a Hill-type curve in clock time; Michaelis-Menten is a special saturating form analogous to a Hill curve with coefficient 1 and lower asymptote 0. Transforming time alone does not provide the mechanistic assumptions or identifiable plateau parameters required by these models.

For a monotone time course that approaches a plateau, an asymptotic exponential model such as `y(t) = plateau - (plateau - y0) * exp(-k * t)` is generally more natural than Michaelis-Menten because it represents first-order accumulation over time. Logistic or Gompertz models are more appropriate only when the data show a genuine lag and sigmoidal transition. With five distinct time points, separate condition-specific lower asymptotes, upper asymptotes, midpoints, and shape parameters would be weakly identified and prone to overfitting. Any such nonlinear model should therefore be a prespecified secondary analysis with few shared parameters, profile or bootstrap confidence intervals, and comparison against the selected empirical models.

## Diagnostic plots

```{r diagnostic-panels, results='asis'}
for (i in seq_len(nrow(plot_endpoint_meta))) {
  cat("\n### ", plot_endpoint_meta$display_endpoint[i], "\n\n", sep = "")
  diagnostic_file <- sub("\\.pdf$", "_diagnostics.pdf", plot_endpoint_meta$plot_file[i])
  cat("![](", file.path("diagnostic_plots", diagnostic_file), ")\n\n", sep = "")
}
```
