---
title: "Isotope-tracer analysis sensitivity benchmark"
author: "Sudip Das"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  absolute_flux_file: "outputs/isotope_tracer_flux_analysis/flux_absolute_long.csv"
  normalized_flux_file: "outputs/isotope_tracer_flux_analysis/normalized_sensitivity_flux_long.csv"
  output_dir: "outputs/isotope_tracer_sensitivity_benchmark"
  write_outputs: true
---

## Purpose

This report checks whether the main isotope-tracer conclusions are robust to
reasonable analysis choices. It benchmarks:

1. raw-scale versus `log10(value + pseudovalue)` categorical-time models;
2. log10 pseudovalues of `0.001`, `0.01`, and `0.1` for absolute abundance;
3. absolute `pmol/sample` versus the optional total-sphingolipid-normalized
   workbook block; and
4. the predefined Pal-d3 and d7 isotope-pooling rules versus their labelled
   component species.

The benchmark evaluates agreement in effect direction and BH-FDR conclusions.
It does not require different response definitions to produce identical
p-values. Fractional-incorporation kinetics and absolute labelled abundance
remain complementary quantities and are not treated as interchangeable.

## Setup

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("dplyr", "tidyr", "ggplot2", "emmeans", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install required packages: ", paste(missing_packages, collapse = ", "))
}

library(dplyr)
library(tidyr)
library(ggplot2)

env_or_default <- function(name, current) {
  value <- Sys.getenv(name, unset = "")
  if (identical(value, "")) current else value
}

absolute_flux_file <- env_or_default(
  "ISOTOPE_BENCHMARK_ABSOLUTE_FILE", params$absolute_flux_file
)
normalized_flux_file <- env_or_default(
  "ISOTOPE_BENCHMARK_NORMALIZED_FILE", params$normalized_flux_file
)
output_dir <- env_or_default(
  "ISOTOPE_BENCHMARK_OUTPUT_DIR", params$output_dir
)
write_outputs <- isTRUE(params$write_outputs)

if (!file.exists(absolute_flux_file)) {
  stop(
    "Missing absolute flux table: ", absolute_flux_file,
    ". Run scripts/isotope_tracer_flux_analysis.Rmd first."
  )
}

if (write_outputs) {
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
}

write_csv <- function(x, filename) {
  if (write_outputs) {
    write.csv(x, file.path(output_dir, filename), row.names = FALSE)
  }
  invisible(x)
}

save_plot <- function(x, filename, width = 8, height = 5) {
  if (write_outputs) {
    ggsave(
      file.path(output_dir, filename), x,
      width = width, height = height, units = "in", bg = "white"
    )
  }
  invisible(x)
}

read_nonempty_csv <- function(path) {
  if (!file.exists(path) || file.info(path)$size == 0) {
    return(tibble())
  }
  tryCatch(
    as_tibble(read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)),
    error = function(e) tibble()
  )
}

times <- c(0.5, 1, 2, 4, 8)
classes <- c("dhCer", "Cer", "dhSM", "SM", "HexCer", "LacCer")
pal_components <- c("d3-16:0", "d6-16:0")
d7_components <- c("d7-16:0", "d7-24:0", "d7-24:1")
lcb_by_experiment <- list(
  `Pal-d3` = c("d3-dhSph", "d3-Sph"),
  `d7-dhSph` = c("d7-dhSph", "d7-Sph", "d7-S1P")
)
```

## Input validation and endpoint construction

```{r input-and-endpoints}
absolute_flux <- read_nonempty_csv(absolute_flux_file)
normalized_flux <- read_nonempty_csv(normalized_flux_file)

required_columns <- c(
  "experiment", "unit", "berlin_sample_id", "infection",
  "is_tracer_treated", "time_h", "replicate", "lipid_class_short",
  "species", "value", "is_class_total", "is_global_total"
)

validate_flux <- function(x, label) {
  if (!nrow(x)) return(invisible(FALSE))
  missing <- setdiff(required_columns, names(x))
  if (length(missing)) {
    stop(label, " is missing columns: ", paste(missing, collapse = ", "))
  }
  if (any(!is.finite(x$value), na.rm = TRUE) || any(x$value < 0, na.rm = TRUE)) {
    stop(label, " contains non-finite or negative measurement values.")
  }
  invisible(TRUE)
}

validate_flux(absolute_flux, "Absolute flux table")
validate_flux(normalized_flux, "Normalized flux table")

filter_analysis_rows <- function(x) {
  x |>
    filter(
      is_tracer_treated,
      !is_class_total,
      !is_global_total,
      infection %in% c("Uninfected", "Infected"),
      time_h %in% times
    ) |>
    mutate(
      infection = factor(infection, levels = c("Uninfected", "Infected")),
      replicate = as.character(replicate)
    )
}

make_lcb_readouts <- function(x) {
  bind_rows(lapply(names(lcb_by_experiment), function(experiment_name) {
    species_keep <- lcb_by_experiment[[experiment_name]]
    x |>
      filter(
        experiment == experiment_name,
        lipid_class_short == "LCB",
        species %in% species_keep
      ) |>
      transmute(
        experiment, unit, berlin_sample_id, infection, time_h, replicate,
        lipid_class_short = "LCB",
        endpoint = species,
        definition = "primary_pooled",
        value
      )
  }))
}

make_pal_class_readouts <- function(x) {
  wide <- x |>
    filter(
      experiment == "Pal-d3",
      lipid_class_short %in% classes,
      species %in% pal_components
    ) |>
    select(
      experiment, unit, berlin_sample_id, infection, time_h, replicate,
      lipid_class_short, species, value
    ) |>
    pivot_wider(names_from = species, values_from = value)

  if (!nrow(wide)) return(tibble())
  if (anyNA(wide[pal_components])) {
    stop("Pal-d3 benchmark is missing a paired d3 or d6 component.")
  }

  bind_rows(
    wide |>
      transmute(
        experiment, unit, berlin_sample_id, infection, time_h, replicate,
        lipid_class_short,
        endpoint = paste(lipid_class_short, "d3+d6-16:0"),
        definition = "primary_pooled",
        value = `d3-16:0` + `d6-16:0`
      ),
    wide |>
      transmute(
        experiment, unit, berlin_sample_id, infection, time_h, replicate,
        lipid_class_short,
        endpoint = paste(lipid_class_short, "d3-only-16:0"),
        definition = "d3_only",
        value = `d3-16:0`
      )
  )
}

make_d7_class_readouts <- function(x) {
  wide <- x |>
    filter(
      experiment == "d7-dhSph",
      lipid_class_short %in% classes,
      species %in% d7_components
    ) |>
    select(
      experiment, unit, berlin_sample_id, infection, time_h, replicate,
      lipid_class_short, species, value
    ) |>
    pivot_wider(names_from = species, values_from = value)

  if (!nrow(wide)) return(tibble())
  if (anyNA(wide[d7_components])) {
    stop("d7 benchmark is missing one or more pooled labelled components.")
  }

  bind_rows(
    wide |>
      transmute(
        experiment, unit, berlin_sample_id, infection, time_h, replicate,
        lipid_class_short,
        endpoint = paste0("d7-", lipid_class_short),
        definition = "primary_pooled",
        value = `d7-16:0` + `d7-24:0` + `d7-24:1`
      ),
    bind_rows(lapply(d7_components, function(component) {
      wide |>
        transmute(
          experiment, unit, berlin_sample_id, infection, time_h, replicate,
          lipid_class_short,
          endpoint = paste(lipid_class_short, component),
          definition = paste0(gsub("[^A-Za-z0-9]+", "_", component), "_only"),
          value = .data[[component]]
        )
    }))
  )
}

make_benchmark_readouts <- function(x) {
  if (!nrow(x)) return(tibble())
  filtered <- filter_analysis_rows(x)
  bind_rows(
    make_lcb_readouts(filtered),
    make_pal_class_readouts(filtered),
    make_d7_class_readouts(filtered)
  ) |>
    arrange(experiment, definition, endpoint, time_h, infection, replicate)
}

absolute_readouts <- make_benchmark_readouts(absolute_flux)
normalized_readouts <- make_benchmark_readouts(normalized_flux)

primary_design <- absolute_readouts |>
  filter(definition == "primary_pooled") |>
  count(experiment, endpoint, time_h, infection, name = "n")

expected_endpoints <- c(`Pal-d3` = 8L, `d7-dhSph` = 9L)
observed_endpoints <- absolute_readouts |>
  filter(definition == "primary_pooled") |>
  distinct(experiment, endpoint) |>
  count(experiment, name = "n_endpoints")

for (experiment_name in names(expected_endpoints)) {
  observed <- observed_endpoints$n_endpoints[
    observed_endpoints$experiment == experiment_name
  ]
  if (length(observed) != 1 || observed != expected_endpoints[[experiment_name]]) {
    stop("Unexpected primary endpoint count for ", experiment_name, ".")
  }
}
if (any(primary_design$n != 3)) {
  stop("Expected three replicates per primary endpoint, time, and infection group.")
}

input_qc <- bind_rows(
  absolute_readouts |>
    summarise(
      source = "absolute",
      available = TRUE,
      rows = n(),
      endpoints = n_distinct(endpoint),
      units = paste(sort(unique(unit)), collapse = ";")
    ),
  if (nrow(normalized_readouts)) {
    normalized_readouts |>
      summarise(
        source = "normalized",
        available = TRUE,
        rows = n(),
        endpoints = n_distinct(endpoint),
        units = paste(sort(unique(unit)), collapse = ";")
      )
  } else {
    tibble(
      source = "normalized", available = FALSE, rows = 0L,
      endpoints = 0L, units = NA_character_
    )
  }
)

write_csv(input_qc, "benchmark_input_qc.csv")
input_qc
```

## Model engine

All timepoint contrasts come from the corresponding interaction model:

```text
response ~ infection * categorical_time
```

Curve FDR is BH-adjusted across the endpoints in the same experiment,
response scale, transformation, and endpoint-definition family. Timepoint FDR
is BH-adjusted across all endpoint-by-time contrasts in that family.

```{r model-engine}
fit_curve <- function(data, transform, pseudovalue) {
  model_data <- data |>
    mutate(
      infection = factor(infection, levels = c("Uninfected", "Infected")),
      time_f = factor(time_h, levels = times),
      response = if (transform == "raw") value else log10(value + pseudovalue)
    )
  fit <- lm(response ~ infection * time_f, data = model_data)
  residual_deviance <- deviance(fit)
  response_scale <- max(1, sum(model_data$response^2, na.rm = TRUE))
  perfect_fit <- residual_deviance <= 100 * .Machine$double.eps * response_scale
  if (perfect_fit) {
    return(tibble(
      n = nrow(model_data),
      p_infection = NA_real_,
      p_time = NA_real_,
      p_interaction = NA_real_,
      shapiro_residual_p = NA_real_,
      max_abs_standardized_residual = NA_real_,
      model_status = "zero_or_near_zero_residual_variance"
    ))
  }
  a <- anova(fit)
  tibble(
    n = nrow(model_data),
    p_infection = a["infection", "Pr(>F)"],
    p_time = a["time_f", "Pr(>F)"],
    p_interaction = a["infection:time_f", "Pr(>F)"],
    shapiro_residual_p = shapiro.test(residuals(fit))$p.value,
    max_abs_standardized_residual = max(abs(rstandard(fit)), na.rm = TRUE),
    model_status = "estimated"
  )
}

fit_timepoints <- function(data, transform, pseudovalue) {
  model_data <- data |>
    mutate(
      infection = factor(infection, levels = c("Uninfected", "Infected")),
      time_f = factor(time_h, levels = times),
      response = if (transform == "raw") value else log10(value + pseudovalue)
    )
  fit <- lm(response ~ infection * time_f, data = model_data)
  residual_deviance <- deviance(fit)
  response_scale <- max(1, sum(model_data$response^2, na.rm = TRUE))
  perfect_fit <- residual_deviance <= 100 * .Machine$double.eps * response_scale
  if (perfect_fit) {
    return(
      model_data |>
        group_by(time_h) |>
        summarise(
          estimate = mean(response[infection == "Infected"]) -
            mean(response[infection == "Uninfected"]),
          .groups = "drop"
        ) |>
        mutate(
          se = NA_real_,
          df = NA_real_,
          t_ratio = NA_real_,
          lower_cl = NA_real_,
          upper_cl = NA_real_,
          p_value = NA_real_,
          direction = case_when(
            estimate > 0 ~ "infected_higher",
            estimate < 0 ~ "infected_lower",
            TRUE ~ "equal"
          ),
          ratio_infected_over_uninfected = if (
            transform == "log10"
          ) 10^estimate else NA_real_,
          model_status = "zero_or_near_zero_residual_variance"
        )
    )
  }
  emmeans::emmeans(fit, ~ infection | time_f) |>
    emmeans::contrast(
      method = list("Infected - Uninfected" = c(-1, 1)),
      adjust = "none"
    ) |>
    summary(infer = c(TRUE, TRUE)) |>
    as_tibble() |>
    transmute(
      time_h = as.numeric(as.character(time_f)),
      estimate,
      se = SE,
      df,
      t_ratio = t.ratio,
      lower_cl = lower.CL,
      upper_cl = upper.CL,
      p_value = p.value,
      direction = case_when(
        estimate > 0 ~ "infected_higher",
        estimate < 0 ~ "infected_lower",
        TRUE ~ "equal"
      ),
      ratio_infected_over_uninfected = if (
        transform == "log10"
      ) 10^estimate else NA_real_,
      model_status = "estimated"
    )
}

run_model_family <- function(data, analysis_id, transform, pseudovalue) {
  if (!nrow(data)) return(list(curves = tibble(), timepoints = tibble()))
  transform_name <- transform
  pseudovalue_used <- pseudovalue

  curves <- data |>
    group_by(
      experiment, unit, definition, lipid_class_short, endpoint
    ) |>
    group_modify(~ fit_curve(.x, transform, pseudovalue)) |>
    ungroup() |>
    mutate(
      analysis_id = analysis_id,
      transform = transform_name,
      pseudovalue = if (
        transform_name == "log10"
      ) pseudovalue_used else NA_real_
    ) |>
    group_by(experiment, unit, definition, analysis_id) |>
    mutate(
      fdr_infection = p.adjust(p_infection, method = "BH"),
      fdr_time = p.adjust(p_time, method = "BH"),
      curve_fdr = p.adjust(p_interaction, method = "BH"),
      significant_curve_fdr_0_05 = curve_fdr < 0.05
    ) |>
    ungroup()

  timepoints <- data |>
    group_by(
      experiment, unit, definition, lipid_class_short, endpoint
    ) |>
    group_modify(~ fit_timepoints(.x, transform, pseudovalue)) |>
    ungroup() |>
    mutate(
      analysis_id = analysis_id,
      transform = transform_name,
      pseudovalue = if (
        transform_name == "log10"
      ) pseudovalue_used else NA_real_
    ) |>
    group_by(experiment, unit, definition, analysis_id) |>
    mutate(
      timepoint_fdr = p.adjust(p_value, method = "BH"),
      significant_timepoint_fdr_0_05 = timepoint_fdr < 0.05
    ) |>
    ungroup()

  list(curves = curves, timepoints = timepoints)
}

absolute_primary <- absolute_readouts |>
  filter(definition == "primary_pooled")

model_runs <- list(
  raw = run_model_family(absolute_primary, "absolute_raw", "raw", NA_real_),
  pv_0001 = run_model_family(
    absolute_primary, "absolute_log10_pv_0.001", "log10", 0.001
  ),
  pv_001 = run_model_family(
    absolute_primary, "absolute_log10_pv_0.01", "log10", 0.01
  ),
  pv_01 = run_model_family(
    absolute_primary, "absolute_log10_pv_0.1", "log10", 0.1
  ),
  pooled_definitions = run_model_family(
    absolute_readouts, "absolute_log10_pooling_definitions", "log10", 0.01
  )
)

if (nrow(normalized_readouts)) {
  model_runs$normalized <- run_model_family(
    normalized_readouts |>
      filter(definition == "primary_pooled"),
    "normalized_log10_pv_0.01", "log10", 0.01
  )
}

curve_models <- bind_rows(lapply(model_runs, `[[`, "curves")) |>
  distinct()
timepoint_models <- bind_rows(lapply(model_runs, `[[`, "timepoints")) |>
  distinct()

write_csv(curve_models, "benchmark_curve_models.csv")
write_csv(timepoint_models, "benchmark_timepoint_contrasts.csv")
```

## Raw versus log10 models

```{r raw-vs-log}
raw_curve <- curve_models |>
  filter(analysis_id == "absolute_raw") |>
  select(
    experiment, lipid_class_short, endpoint,
    raw_curve_fdr = curve_fdr,
    raw_significant = significant_curve_fdr_0_05
  )

log_curve <- curve_models |>
  filter(analysis_id == "absolute_log10_pv_0.01") |>
  select(
    experiment, lipid_class_short, endpoint,
    log10_curve_fdr = curve_fdr,
    log10_significant = significant_curve_fdr_0_05
  )

raw_vs_log_curve <- inner_join(
  raw_curve, log_curve,
  by = c("experiment", "lipid_class_short", "endpoint")
) |>
  mutate(significance_agreement = raw_significant == log10_significant)

raw_time <- timepoint_models |>
  filter(analysis_id == "absolute_raw") |>
  select(
    experiment, lipid_class_short, endpoint, time_h,
    raw_direction = direction,
    raw_timepoint_fdr = timepoint_fdr,
    raw_significant = significant_timepoint_fdr_0_05
  )

log_time <- timepoint_models |>
  filter(analysis_id == "absolute_log10_pv_0.01") |>
  select(
    experiment, lipid_class_short, endpoint, time_h,
    log10_direction = direction,
    log10_timepoint_fdr = timepoint_fdr,
    log10_significant = significant_timepoint_fdr_0_05
  )

raw_vs_log_timepoint <- inner_join(
  raw_time, log_time,
  by = c("experiment", "lipid_class_short", "endpoint", "time_h")
) |>
  mutate(
    direction_agreement = raw_direction == log10_direction,
    significance_agreement = raw_significant == log10_significant
  )

write_csv(raw_vs_log_curve, "raw_vs_log_curve_agreement.csv")
write_csv(raw_vs_log_timepoint, "raw_vs_log_timepoint_agreement.csv")

p_raw_log <- ggplot(
  raw_vs_log_curve,
  aes(
    -log10(pmax(raw_curve_fdr, 1e-12)),
    -log10(pmax(log10_curve_fdr, 1e-12)),
    colour = experiment
  )
) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(size = 3) +
  labs(
    x = "Raw-model -log10(curve FDR)",
    y = "Log10-model -log10(curve FDR)",
    colour = "Experiment"
  ) +
  theme_classic(base_size = 11)

save_plot(p_raw_log, "raw_vs_log_curve_fdr.pdf")
raw_vs_log_curve
```

## Pseudovalue sensitivity

```{r pseudovalue-sensitivity}
pseudovalue_curves <- curve_models |>
  filter(
    analysis_id %in% c(
      "absolute_log10_pv_0.001",
      "absolute_log10_pv_0.01",
      "absolute_log10_pv_0.1"
    )
  ) |>
  select(
    experiment, lipid_class_short, endpoint, pseudovalue,
    curve_fdr, significant_curve_fdr_0_05
  )

pseudovalue_curve_stability <- pseudovalue_curves |>
  group_by(experiment, lipid_class_short, endpoint) |>
  summarise(
    n_estimable = sum(!is.na(curve_fdr)),
    min_curve_fdr = if (n_estimable > 0) min(curve_fdr, na.rm = TRUE) else NA_real_,
    max_curve_fdr = if (n_estimable > 0) max(curve_fdr, na.rm = TRUE) else NA_real_,
    significance_stable = n_estimable == 3 &&
      n_distinct(significant_curve_fdr_0_05[!is.na(curve_fdr)]) == 1,
    .groups = "drop"
  )

pseudovalue_timepoint_stability <- timepoint_models |>
  filter(
    analysis_id %in% c(
      "absolute_log10_pv_0.001",
      "absolute_log10_pv_0.01",
      "absolute_log10_pv_0.1"
    )
  ) |>
  group_by(experiment, lipid_class_short, endpoint, time_h) |>
  summarise(
    n_estimable = sum(!is.na(timepoint_fdr)),
    direction_stable = n_distinct(direction) == 1,
    significance_stable = n_estimable == 3 &&
      n_distinct(significant_timepoint_fdr_0_05[!is.na(timepoint_fdr)]) == 1,
    min_timepoint_fdr = if (
      n_estimable > 0
    ) min(timepoint_fdr, na.rm = TRUE) else NA_real_,
    max_timepoint_fdr = if (
      n_estimable > 0
    ) max(timepoint_fdr, na.rm = TRUE) else NA_real_,
    .groups = "drop"
  )

write_csv(pseudovalue_curves, "pseudovalue_curve_results.csv")
write_csv(
  pseudovalue_curve_stability,
  "pseudovalue_curve_stability.csv"
)
write_csv(
  pseudovalue_timepoint_stability,
  "pseudovalue_timepoint_stability.csv"
)

p_pseudovalue <- ggplot(
  pseudovalue_curves,
  aes(factor(pseudovalue), -log10(pmax(curve_fdr, 1e-12)), group = endpoint)
) +
  geom_line(colour = "grey55", linewidth = 0.5) +
  geom_point(aes(colour = experiment), size = 2.2) +
  facet_wrap(~ experiment, scales = "free_y") +
  labs(
    x = "Log10 pseudovalue",
    y = "-log10(curve FDR)",
    colour = "Experiment"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "none")

save_plot(p_pseudovalue, "pseudovalue_curve_fdr_sensitivity.pdf")
pseudovalue_curve_stability
```

## Absolute versus normalized block

This comparison is generated only when the normalized workbook block was
imported by `scripts/isotope_tracer_flux_analysis.Rmd`. Normalization can alter
the scientific estimand, so agreement supports robustness but disagreement is
not automatically evidence that either scale is invalid.

```{r absolute-vs-normalized}
if ("normalized" %in% names(model_runs)) {
  absolute_curve <- curve_models |>
    filter(analysis_id == "absolute_log10_pv_0.01") |>
    select(
      experiment, lipid_class_short, endpoint,
      absolute_curve_fdr = curve_fdr,
      absolute_significant = significant_curve_fdr_0_05
    )
  normalized_curve <- curve_models |>
    filter(analysis_id == "normalized_log10_pv_0.01") |>
    select(
      experiment, lipid_class_short, endpoint,
      normalized_curve_fdr = curve_fdr,
      normalized_significant = significant_curve_fdr_0_05
    )
  absolute_vs_normalized_curve <- inner_join(
    absolute_curve, normalized_curve,
    by = c("experiment", "lipid_class_short", "endpoint")
  ) |>
    mutate(
      significance_agreement = absolute_significant == normalized_significant
    )

  absolute_time <- timepoint_models |>
    filter(analysis_id == "absolute_log10_pv_0.01") |>
    select(
      experiment, lipid_class_short, endpoint, time_h,
      absolute_direction = direction,
      absolute_timepoint_fdr = timepoint_fdr,
      absolute_significant = significant_timepoint_fdr_0_05
    )
  normalized_time <- timepoint_models |>
    filter(analysis_id == "normalized_log10_pv_0.01") |>
    select(
      experiment, lipid_class_short, endpoint, time_h,
      normalized_direction = direction,
      normalized_timepoint_fdr = timepoint_fdr,
      normalized_significant = significant_timepoint_fdr_0_05
    )
  absolute_vs_normalized_timepoint <- inner_join(
    absolute_time, normalized_time,
    by = c("experiment", "lipid_class_short", "endpoint", "time_h")
  ) |>
    mutate(
      direction_agreement = absolute_direction == normalized_direction,
      significance_agreement = absolute_significant == normalized_significant
    )

  p_absolute_normalized <- ggplot(
    absolute_vs_normalized_curve,
    aes(
      -log10(pmax(absolute_curve_fdr, 1e-12)),
      -log10(pmax(normalized_curve_fdr, 1e-12)),
      colour = experiment
    )
  ) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
    geom_point(size = 3) +
    labs(
      x = "Absolute -log10(curve FDR)",
      y = "Normalized-block -log10(curve FDR)",
      colour = "Experiment"
    ) +
    theme_classic(base_size = 11)

  write_csv(
    absolute_vs_normalized_curve,
    "absolute_vs_normalized_curve_comparison.csv"
  )
  write_csv(
    absolute_vs_normalized_timepoint,
    "absolute_vs_normalized_timepoint_comparison.csv"
  )
  save_plot(
    p_absolute_normalized,
    "absolute_vs_normalized_curve_fdr.pdf"
  )
} else {
  absolute_vs_normalized_curve <- tibble()
  absolute_vs_normalized_timepoint <- tibble()
}

absolute_vs_normalized_curve
```

## Isotope-pooling sensitivity

The primary Pal-d3 endpoint sums d3 and d6 for the relevant 16:0 species. The
primary d7 class endpoint sums d7-16:0, d7-24:0, and d7-24:1. Comparisons with
single components are sensitivity analyses of endpoint construction, not
tests of identical biological quantities. LCB molecules remain individual and
are therefore excluded from this pooling comparison.

```{r pooling-sensitivity}
pool_curves <- curve_models |>
  filter(
    analysis_id == "absolute_log10_pooling_definitions",
    lipid_class_short != "LCB"
  ) |>
  select(
    experiment, lipid_class_short, definition, endpoint,
    curve_fdr, significant_curve_fdr_0_05
  )

primary_pool_curves <- pool_curves |>
  filter(definition == "primary_pooled") |>
  select(
    experiment, lipid_class_short,
    primary_endpoint = endpoint,
    primary_curve_fdr = curve_fdr,
    primary_significant = significant_curve_fdr_0_05
  )

pooling_curve_comparison <- pool_curves |>
  filter(definition != "primary_pooled") |>
  left_join(primary_pool_curves, by = c("experiment", "lipid_class_short")) |>
  rename(
    component_endpoint = endpoint,
    component_curve_fdr = curve_fdr,
    component_significant = significant_curve_fdr_0_05
  ) |>
  mutate(
    significance_agreement = component_significant == primary_significant
  )

pool_timepoints <- timepoint_models |>
  filter(
    analysis_id == "absolute_log10_pooling_definitions",
    lipid_class_short != "LCB"
  )

primary_pool_timepoints <- pool_timepoints |>
  filter(definition == "primary_pooled") |>
  select(
    experiment, lipid_class_short, time_h,
    primary_endpoint = endpoint,
    primary_direction = direction,
    primary_timepoint_fdr = timepoint_fdr,
    primary_significant = significant_timepoint_fdr_0_05
  )

pooling_timepoint_comparison <- pool_timepoints |>
  filter(definition != "primary_pooled") |>
  left_join(
    primary_pool_timepoints,
    by = c("experiment", "lipid_class_short", "time_h")
  ) |>
  rename(
    component_endpoint = endpoint,
    component_direction = direction,
    component_timepoint_fdr = timepoint_fdr,
    component_significant = significant_timepoint_fdr_0_05
  ) |>
  mutate(
    direction_agreement = component_direction == primary_direction,
    significance_agreement = component_significant == primary_significant
  )

write_csv(pooling_curve_comparison, "pooling_curve_comparison.csv")
write_csv(pooling_timepoint_comparison, "pooling_timepoint_comparison.csv")
pooling_curve_comparison
```

## Benchmark summary

```{r benchmark-summary}
agreement_row <- function(comparison, data, direction_col = NULL) {
  direction_values <- if (
    !is.null(direction_col) && direction_col %in% names(data)
  ) data[[direction_col]] else logical()
  significance_values <- if (
    "significance_agreement" %in% names(data)
  ) data$significance_agreement else logical()
  tibble(
    comparison = comparison,
    n_direction_compared = sum(!is.na(direction_values)),
    n_fdr_compared = sum(!is.na(significance_values)),
    direction_agreement_pct = if (sum(!is.na(direction_values)) > 0) {
      100 * mean(direction_values, na.rm = TRUE)
    } else {
      NA_real_
    },
    significance_agreement_pct = if (sum(!is.na(significance_values)) > 0) {
      100 * mean(significance_values, na.rm = TRUE)
    } else {
      NA_real_
    }
  )
}

benchmark_overview <- bind_rows(
  agreement_row("Raw vs log10: curve FDR", raw_vs_log_curve),
  agreement_row(
    "Raw vs log10: timepoint direction/FDR",
    raw_vs_log_timepoint,
    "direction_agreement"
  ),
  tibble(
    comparison = "Pseudovalue: curve FDR stability",
    n_direction_compared = 0L,
    n_fdr_compared = sum(pseudovalue_curve_stability$n_estimable == 3),
    direction_agreement_pct = NA_real_,
    significance_agreement_pct = 100 * mean(
      pseudovalue_curve_stability$significance_stable[
        pseudovalue_curve_stability$n_estimable == 3
      ],
      na.rm = TRUE
    )
  ),
  tibble(
    comparison = "Pseudovalue: timepoint direction/FDR stability",
    n_direction_compared = nrow(pseudovalue_timepoint_stability),
    n_fdr_compared = sum(pseudovalue_timepoint_stability$n_estimable == 3),
    direction_agreement_pct = 100 * mean(
      pseudovalue_timepoint_stability$direction_stable,
      na.rm = TRUE
    ),
    significance_agreement_pct = 100 * mean(
      pseudovalue_timepoint_stability$significance_stable[
        pseudovalue_timepoint_stability$n_estimable == 3
      ],
      na.rm = TRUE
    )
  ),
  agreement_row(
    "Absolute vs normalized: curve FDR",
    absolute_vs_normalized_curve
  ),
  agreement_row(
    "Absolute vs normalized: timepoint direction/FDR",
    absolute_vs_normalized_timepoint,
    "direction_agreement"
  ),
  agreement_row(
    "Pooled vs component: curve FDR",
    pooling_curve_comparison
  ),
  agreement_row(
    "Pooled vs component: timepoint direction/FDR",
    pooling_timepoint_comparison,
    "direction_agreement"
  )
)

write_csv(benchmark_overview, "benchmark_overview.csv")

format_pct <- function(x) {
  ifelse(is.na(x), "not available", sprintf("%.1f%%", x))
}

summary_lines <- c(
  "# Isotope-Tracer Sensitivity Benchmark",
  "",
  "The primary analysis uses absolute `pmol/sample`, predefined isotope pooling, and `log10(value + 0.01)` categorical-time interaction models.",
  "",
  "## Agreement Summary",
  "",
  "| Comparison | n direction | n FDR | Direction agreement | FDR-decision agreement |",
  "| --- | ---: | ---: | ---: | ---: |",
  vapply(seq_len(nrow(benchmark_overview)), function(i) {
    paste0(
      "| ", benchmark_overview$comparison[i],
      " | ", benchmark_overview$n_direction_compared[i],
      " | ", benchmark_overview$n_fdr_compared[i],
      " | ", format_pct(benchmark_overview$direction_agreement_pct[i]),
      " | ", format_pct(benchmark_overview$significance_agreement_pct[i]),
      " |"
    )
  }, character(1)),
  "",
  "## Interpretation Limits",
  "",
  "- Agreement means that conclusions are not highly sensitive to the tested choice; it does not prove that every model is equally appropriate.",
  "- Absolute and total-sphingolipid-normalized blocks answer different quantitative questions.",
  "- Pooled and single-component isotope endpoints are related but non-identical biological quantities.",
  "- Fractional-incorporation kinetics should be interpreted beside, not substituted for, absolute labelled abundance.",
  "- FDR < 0.05 is used as the decision threshold, but effect direction, uncertainty, diagnostics, and biological context remain important."
)

if (write_outputs) {
  writeLines(summary_lines, file.path(output_dir, "BENCHMARK_SUMMARY.md"))
}

knitr::kable(benchmark_overview, digits = 1)
```
