---
title: "Isotope Tracer Flux Analysis"
author: "Sudip Das"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  pal_d3_input_file: ""
  d7_dhsph_input_file: ""
  output_dir: "outputs/isotope_tracer_flux_analysis"
  pseudovalue: 0.01
  include_normalized_block: false
  export_individual_panels: false
  individual_panel_formats: "png,pdf"
  write_outputs: false
---

## 1. Purpose

This notebook starts the isotope-tracing analysis for the Pal-d3 and
d7-dhSph flux experiments.

The main goal is to create one adaptable tidy data structure from both
workbooks. The downstream analysis should then use the same columns for both
tracers:

- `flux_long`: one row per sample and measured lipid column
- `flux_absolute_long`: absolute `pmol / sample` rows only
- `flux_species_readouts`: non-LCB species-level labelled and unlabelled
  readouts
- `flux_class_readouts`: supporting non-LCB class-level labelled, unlabelled,
  and workbook-total readouts; these are not used as the primary tracer
  analysis features
- `flux_lcb_readouts`: molecule-level LCB labelled and unlabelled readouts
- `flux_analysis_readouts`: combined primary analysis table with non-LCB
  species plus molecule-level LCB features
- `flux_labelled_species`: individual labelled species, useful for inspecting
  d7-dhSph conversion products such as `d7-Sph` and `d7-S1P`
The absolute `pmol / sample` block is the primary analysis block. The normalized
`fmol / pmol total sphingolipids` block can be imported by setting
`params$include_normalized_block = TRUE`, but it is not used as the default
primary analysis.

## 2. Setup

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(emmeans)

`%||%` <- function(x, y) {
  if (is.null(x)) {
    y
  } else {
    x
  }
}

if (!exists("params")) {
  params <- list()
}

params$pal_d3_input_file <- params$pal_d3_input_file %||% ""
params$d7_dhsph_input_file <- params$d7_dhsph_input_file %||% ""
params$output_dir <- params$output_dir %||% "outputs/isotope_tracer_flux_analysis"
params$pseudovalue <- params$pseudovalue %||% 0.01
params$include_normalized_block <- params$include_normalized_block %||% FALSE
params$export_individual_panels <- params$export_individual_panels %||% FALSE
params$individual_panel_formats <- params$individual_panel_formats %||% "png"
params$write_outputs <- params$write_outputs %||% FALSE

env_or_default <- function(name, current) {
  value <- Sys.getenv(name, unset = "")
  if (identical(value, "")) current else value
}

params$pal_d3_input_file <- env_or_default(
  "PAL_D3_INPUT_FILE", params$pal_d3_input_file
)
params$d7_dhsph_input_file <- env_or_default(
  "D7_DHSPH_INPUT_FILE", params$d7_dhsph_input_file
)
params$output_dir <- env_or_default(
  "ISOTOPE_OUTPUT_DIR", params$output_dir
)
params$pseudovalue <- as.numeric(env_or_default(
  "ISOTOPE_LOG10_PSEUDOVALUE", params$pseudovalue
))

write_outputs_env <- Sys.getenv("WRITE_OUTPUTS", unset = "")
write_outputs <- isTRUE(params$write_outputs) ||
  tolower(write_outputs_env) %in% c("true", "1", "yes")

include_normalized_block_env <- Sys.getenv("INCLUDE_NORMALIZED_BLOCK", unset = "")
include_normalized_block <- isTRUE(params$include_normalized_block) ||
  tolower(include_normalized_block_env) %in% c("true", "1", "yes")

export_individual_panels_env <- Sys.getenv("EXPORT_INDIVIDUAL_PANELS", unset = "")
export_individual_panels <- isTRUE(params$export_individual_panels) ||
  tolower(export_individual_panels_env) %in% c("true", "1", "yes")

individual_panel_formats_env <- Sys.getenv("INDIVIDUAL_PANEL_FORMATS", unset = "")
individual_panel_formats <- if (!identical(individual_panel_formats_env, "")) {
  individual_panel_formats_env
} else {
  params$individual_panel_formats
}

individual_panel_formats <- individual_panel_formats |>
  str_split(",") |>
  unlist() |>
  str_squish() |>
  str_to_lower()

individual_panel_formats <- individual_panel_formats[
  individual_panel_formats %in% c("png", "pdf")
]

if (length(individual_panel_formats) == 0) {
  stop("individual_panel_formats must include png, pdf, or both.")
}

output_dir <- params$output_dir
pseudovalue <- as.numeric(params$pseudovalue)

if (!is.finite(pseudovalue) || pseudovalue <= 0) {
  stop("params$pseudovalue must be a positive number.")
}

non_lcb_lipid_classes <- c("dhCer", "Cer", "dhSM", "SM", "HexCer", "LacCer")
non_lcb_species_levels <- c(
  "16:0", "18:0", "20:0", "22:0", "24:0", "24:1",
  "d3-16:0", "d6-16:0",
  "d7-16:0", "d7-24:0", "d7-24:1"
)
lcb_species_levels <- c(
  "dhSph", "d3-dhSph", "d7-dhSph",
  "Sph", "d3-Sph", "d7-Sph",
  "S1P", "d7-S1P"
)
analysis_feature_type_levels <- c("LCB molecule", "Lipid species")
timepoint_levels <- c(0.5, 1, 2, 4, 8)
timepoint_labels <- c("0.5", "1", "2", "4", "8")
timepoint_plot_breaks <- seq_along(timepoint_levels)

format_analysis_feature <- function(lipid_class_short, species) {
  ifelse(
    as.character(lipid_class_short) == "LCB",
    as.character(species),
    paste(as.character(lipid_class_short), as.character(species))
  )
}

species_sort_index <- function(species) {
  species_levels <- c(lcb_species_levels, non_lcb_species_levels)
  index <- match(as.character(species), species_levels)
  ifelse(is.na(index), Inf, index)
}

add_timepoint_plot_position <- function(data) {
  data |>
    mutate(time_plot = match(time_h, timepoint_levels))
}

scale_timepoint_x <- function() {
  scale_x_continuous(
    breaks = timepoint_plot_breaks,
    labels = timepoint_labels
  )
}

if (write_outputs) {
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
}

write_csv_if_requested <- function(x, filename) {
  if (write_outputs) {
    write.csv(x, file.path(output_dir, filename), row.names = FALSE)
  }
  invisible(x)
}

save_plot_if_requested <- function(plot, filename, width = 10, height = 6) {
  if (write_outputs) {
    ggsave(
      filename = file.path(output_dir, filename),
      plot = plot,
      width = width,
      height = height,
      units = "in",
      dpi = 300,
      bg = "white"
    )
  }
  invisible(plot)
}

safe_file_part <- function(x) {
  x |>
    as.character() |>
    str_replace_all("[^A-Za-z0-9]+", "_") |>
    str_replace_all("_+", "_") |>
    str_replace_all("^_|_$", "") |>
    str_to_lower()
}

save_individual_panel_if_requested <- function(plot, path_stem, width = 4, height = 3) {
  if (write_outputs && export_individual_panels) {
    panel_dir <- dirname(file.path(output_dir, path_stem))
    dir.create(panel_dir, recursive = TRUE, showWarnings = FALSE)

    for (format in individual_panel_formats) {
      ggsave(
        filename = file.path(output_dir, paste0(path_stem, ".", format)),
        plot = plot,
        width = width,
        height = height,
        units = "in",
        dpi = 300,
        bg = "white"
      )
    }
  }

  invisible(plot)
}

format_curve_fdr_label <- function(p_value) {
  case_when(
    is.na(p_value) ~ "curve FDR = NA",
    p_value < 0.001 ~ "curve FDR < 0.001",
    TRUE ~ paste0("curve FDR = ", formatC(p_value, format = "f", digits = 3))
  )
}

curve_label_positions <- function(summary_data, group_cols) {
  summary_data |>
    group_by(across(all_of(group_cols))) |>
    summarise(
      label_time = min(time_h, na.rm = TRUE),
      y_min = min(pmax(mean_value - se_value, 0), na.rm = TRUE),
      y_max = max(mean_value + se_value, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      label_x = timepoint_plot_breaks[match(label_time, timepoint_levels)],
      label_x = ifelse(is.na(label_x), min(timepoint_plot_breaks), label_x),
      y_range = ifelse(
        is.finite(y_max - y_min) & y_max > y_min,
        y_max - y_min,
        pmax(abs(y_max), 1)
      ),
      label_y = ifelse(is.finite(y_max), y_max + 0.08 * y_range, NA_real_)
    ) |>
    select(-label_time, -y_min, -y_max, -y_range)
}
```

## 3. Workbook Configuration

The file paths can be supplied through the YAML parameters. If they are left
blank, the notebook searches the project root and `data/` for one workbook
containing `Pal-d3` and one workbook containing `d7-dhSph` in the filename.

```{r workbook-config}
resolve_input_file <- function(explicit_path, pattern, label) {
  if (!identical(explicit_path, "") && !is.na(explicit_path)) {
    if (!file.exists(explicit_path)) {
      stop(label, " input file does not exist: ", explicit_path)
    }
    return(explicit_path)
  }

  candidates <- unlist(lapply(c(".", "data"), function(search_dir) {
    list.files(
      path = search_dir,
      pattern = pattern,
      full.names = TRUE,
      ignore.case = TRUE
    )
  }), use.names = FALSE)
  candidates <- candidates[!grepl("^~[$]", basename(candidates))]

  if (length(candidates) != 1) {
    stop(
      "Set the input file for ", label,
      ". Found candidates: ", paste(candidates, collapse = ", ")
    )
  }

  candidates
}

pal_d3_file <- resolve_input_file(
  explicit_path = params$pal_d3_input_file,
  pattern = "Pal-d3.*[.]xlsx$",
  label = "Pal-d3"
)

d7_dhsph_file <- resolve_input_file(
  explicit_path = params$d7_dhsph_input_file,
  pattern = "d7.*dhSph.*[.]xlsx$",
  label = "d7-dhSph"
)

tracer_specs <- list(
  pal_d3 = list(
    experiment = "Pal-d3",
    input_file = pal_d3_file,
    sheet = "Tabelle1",
    tracer = "Pal-d3",
    tracer_regex = "d3[- ]?Palmitate",
    isotope_regex = "^d[36]-",
    labelled_readout_name = "labelled_flux_d3_plus_d6"
  ),
  d7_dhsph = list(
    experiment = "d7-dhSph",
    input_file = d7_dhsph_file,
    sheet = "d7-dhSph",
    tracer = "d7-dhSph",
    tracer_regex = "d7[- ]?dhSph",
    isotope_regex = "^d7-",
    labelled_readout_name = "labelled_flux_d7"
  )
)

block_specs <- list(
  absolute = list(
    block = "absolute",
    unit = "pmol_per_sample",
    class_row = 3,
    header_row = 4,
    data_start = 5,
    data_end = 64
  ),
  normalized = list(
    block = "normalized",
    unit = "fmol_per_pmol_total_sphingolipids",
    class_row = 69,
    header_row = 70,
    data_start = 71,
    data_end = 130
  )
)

blocks_to_import <- block_specs["absolute"]

if (include_normalized_block) {
  blocks_to_import <- block_specs
}

list(
  pal_d3_file = pal_d3_file,
  d7_dhsph_file = d7_dhsph_file,
  blocks_to_import = names(blocks_to_import),
  pseudovalue = pseudovalue
)
```

## 4. Generic Import Functions

These functions are intentionally tracer-generic. To add another isotope
experiment later, define one additional entry in `tracer_specs` with the
workbook, sheet, tracer-name regex, and isotope-label regex.

```{r import-functions}
clean_label <- function(x) {
  str_squish(as.character(x))
}

fill_right <- function(x) {
  out <- x
  last <- NA_character_

  for (i in seq_along(out)) {
    if (!is.na(out[i]) && out[i] != "NA" && out[i] != "") {
      last <- out[i]
    }
    out[i] <- last
  }

  out
}

make_lipid_id <- function(lipid_class_short, species_base, label_status) {
  species_id <- species_base |>
    str_replace_all(":", "_") |>
    str_replace_all("[()]", "") |>
    str_replace_all("[ /-]+", "_") |>
    str_replace_all("[^A-Za-z0-9_]", "") |>
    str_replace_all("_+", "_") |>
    str_replace_all("^_|_$", "") |>
    str_to_lower()

  class_id <- lipid_class_short |>
    str_replace_all("[^A-Za-z0-9]+", "_") |>
    str_replace_all("_+", "_") |>
    str_replace_all("^_|_$", "") |>
    str_to_lower()

  ifelse(
    lipid_class_short == "LCB",
    paste(label_status, species_id, sep = "_"),
    paste(label_status, class_id, species_id, sep = "_")
  )
}

class_short_name <- function(lipid_class) {
  case_when(
    lipid_class == "LCBs" ~ "LCB",
    lipid_class == "Dihydroceramides (dhCer)" ~ "dhCer",
    lipid_class == "Ceramides (Cer)" ~ "Cer",
    lipid_class == "Dihydrosphingomyelins (dhSM)" ~ "dhSM",
    lipid_class == "Sphingomyelins (SM)" ~ "SM",
    lipid_class == "Hexosylceramides (HexCer)" ~ "HexCer",
    lipid_class == "Lactosylceramides (LacCer)" ~ "LacCer",
    lipid_class == "Total sphingolipid" ~ "Total sphingolipid",
    TRUE ~ lipid_class
  )
}

parse_tracer_block <- function(raw, spec, block) {
  classes <- clean_label(unlist(raw[block$class_row, ]))
  species <- clean_label(unlist(raw[block$header_row, ]))
  classes_filled <- fill_right(classes)

  column_map <- data.frame(
    column_index = seq_along(species),
    lipid_class = classes_filled,
    species = species,
    stringsAsFactors = FALSE
  ) |>
    filter(
      column_index > 3,
      !is.na(species),
      species != "NA",
      species != "",
      !is.na(lipid_class)
    ) |>
    mutate(
      lipid_class = ifelse(
        str_detect(species, regex("total sphingolipid", ignore_case = TRUE)),
        "Total sphingolipid",
        lipid_class
      ),
      lipid_class_short = class_short_name(lipid_class),
      label_status = case_when(
        str_detect(species, "^d3-") ~ "labelled_d3",
        str_detect(species, "^d6-") ~ "labelled_d6",
        str_detect(species, "^d7-") ~ "labelled_d7",
        TRUE ~ "unlabelled"
      ),
      label_group = ifelse(
        str_detect(species, regex(spec$isotope_regex)),
        "labelled",
        "unlabelled"
      ),
      species_base = species |>
        str_replace("^d3-", "") |>
        str_replace("^d6-", "") |>
        str_replace("^d7-", ""),
      is_class_total = str_detect(species, regex(" total$", ignore_case = TRUE)),
      is_global_total = str_detect(
        species,
        regex("total sphingolipid", ignore_case = TRUE)
      ),
      lipid_id = make_lipid_id(lipid_class_short, species_base, label_status)
    )

  data_block <- raw[block$data_start:block$data_end, , drop = FALSE]
  names(data_block) <- paste0("V", seq_len(ncol(data_block)))

  metadata <- tibble(
    experiment = spec$experiment,
    tracer = spec$tracer,
    source_block = block$block,
    unit = block$unit,
    berlin_sample_id = suppressWarnings(as.integer(unlist(data_block[1]))),
    sample_name = clean_label(unlist(data_block[2]))
  ) |>
    mutate(
      time_h = as.numeric(str_match(sample_name, "(0[.]5|1|2|4|8)h")[, 2]),
      replicate = str_match(sample_name, "R([123])")[, 2],
      infection = ifelse(
        str_detect(sample_name, regex("Sne", ignore_case = TRUE)),
        "Infected",
        "Uninfected"
      ),
      tracer_treatment = ifelse(
        str_detect(sample_name, regex(spec$tracer_regex, ignore_case = TRUE)),
        spec$tracer,
        paste("No", spec$tracer)
      ),
      is_tracer_treated = tracer_treatment == spec$tracer,
      sample_group = case_when(
        infection == "Uninfected" & !is_tracer_treated ~ "Uninfected",
        infection == "Uninfected" & is_tracer_treated ~
          paste("Uninfected +", spec$tracer),
        infection == "Infected" & !is_tracer_treated ~ "Infected",
        infection == "Infected" & is_tracer_treated ~
          paste("Infected +", spec$tracer),
        TRUE ~ sample_name
      )
    )

  value_block <- data_block[, column_map$column_index, drop = FALSE]
  names(value_block) <- paste0("C", column_map$column_index)

  bind_cols(metadata, value_block) |>
    pivot_longer(
      cols = starts_with("C"),
      names_to = "column_key",
      values_to = "value"
    ) |>
    mutate(
      column_index = as.integer(str_remove(column_key, "^C")),
      value = suppressWarnings(as.numeric(value))
    ) |>
    left_join(column_map, by = "column_index") |>
    select(
      experiment,
      tracer,
      source_block,
      unit,
      berlin_sample_id,
      sample_name,
      sample_group,
      infection,
      tracer_treatment,
      is_tracer_treated,
      time_h,
      replicate,
      lipid_class,
      lipid_class_short,
      species,
      species_base,
      lipid_id,
      label_status,
      label_group,
      is_class_total,
      is_global_total,
      value
    ) |>
    mutate(
      value_log10 = log10(value + pseudovalue),
      infection = factor(infection, levels = c("Uninfected", "Infected")),
      lipid_class_short = factor(
        lipid_class_short,
        levels = c(
          "LCB", "dhCer", "Cer", "dhSM", "SM", "HexCer", "LacCer",
          "Total sphingolipid"
        )
      )
    )
}

read_tracer_workbook <- function(spec, blocks) {
  raw <- read_excel(
    spec$input_file,
    sheet = spec$sheet,
    col_names = FALSE,
    guess_max = 1000
  )

  block_tables <- lapply(blocks, function(block) {
    parse_tracer_block(raw = raw, spec = spec, block = block)
  })

  bind_rows(block_tables)
}
```

## 5. Create Shared Data Frames

```{r create-data-frames}
flux_long <- bind_rows(lapply(tracer_specs, read_tracer_workbook, blocks = blocks_to_import)) |>
  mutate(
    experiment = factor(experiment, levels = c("Pal-d3", "d7-dhSph"))
  )

flux_absolute_long <- flux_long |>
  filter(unit == "pmol_per_sample")

sample_design <- flux_absolute_long |>
  distinct(
    experiment,
    tracer,
    berlin_sample_id,
    sample_name,
    sample_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h,
    replicate
  ) |>
  arrange(experiment, replicate, time_h, sample_group)

unlabelled_class <- flux_absolute_long |>
  filter(
    lipid_class_short %in% non_lcb_lipid_classes,
    label_group == "unlabelled",
    !is_class_total,
    !is_global_total
  ) |>
  group_by(
    experiment,
    tracer,
    berlin_sample_id,
    sample_name,
    sample_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h,
    replicate,
    unit,
    lipid_class_short
  ) |>
  summarise(
    readout = "unlabelled_abundance",
    readout_label = "Unlabelled abundance",
    value = sum(value, na.rm = TRUE),
    contributing_species = paste(sort(unique(species)), collapse = ";"),
    .groups = "drop"
  )

labelled_class <- flux_absolute_long |>
  filter(
    lipid_class_short %in% non_lcb_lipid_classes,
    label_group == "labelled",
    !is_class_total,
    !is_global_total
  ) |>
  group_by(
    experiment,
    tracer,
    berlin_sample_id,
    sample_name,
    sample_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h,
    replicate,
    unit,
    lipid_class_short
  ) |>
  summarise(
    readout = "labelled_incorporation",
    readout_label = ifelse(
      first(experiment) == "Pal-d3",
      "labelled_flux_d3_plus_d6",
      "labelled_flux_d7"
    ),
    value = sum(value, na.rm = TRUE),
    contributing_species = paste(sort(unique(species)), collapse = ";"),
    .groups = "drop"
  )

class_total <- flux_absolute_long |>
  filter(
    lipid_class_short %in% non_lcb_lipid_classes,
    is_class_total,
    !is_global_total
  ) |>
  transmute(
    experiment,
    tracer,
    berlin_sample_id,
    sample_name,
    sample_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h,
    replicate,
    unit,
    lipid_class_short,
    readout = "class_total_in_workbook",
    readout_label = "Class total in workbook",
    value,
    contributing_species = species
  )

flux_class_readouts <- bind_rows(unlabelled_class, labelled_class, class_total) |>
  mutate(
    pseudovalue = pseudovalue,
    value_log10 = log10(value + pseudovalue),
    readout = factor(
      readout,
      levels = c(
        "labelled_incorporation",
        "unlabelled_abundance",
        "class_total_in_workbook"
      )
    ),
    lipid_class_short = factor(
      lipid_class_short,
      levels = non_lcb_lipid_classes
    )
  ) |>
  arrange(experiment, readout, lipid_class_short, replicate, time_h, sample_group)

flux_lcb_readouts <- flux_absolute_long |>
  filter(
    lipid_class_short == "LCB",
    label_group %in% c("labelled", "unlabelled"),
    !is_class_total,
    !is_global_total
  ) |>
  transmute(
    experiment,
    tracer,
    berlin_sample_id,
    sample_name,
    sample_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h,
    replicate,
    unit,
    lipid_class_short,
    species,
    species_base,
    label_status,
    label_group,
    readout = ifelse(
      label_group == "labelled",
      "labelled_incorporation",
      "unlabelled_abundance"
    ),
    readout_label = case_when(
      label_group == "labelled" & experiment == "Pal-d3" ~ "labelled_LCB_d3",
      label_group == "labelled" & experiment == "d7-dhSph" ~ "labelled_LCB_d7",
      TRUE ~ "Unlabelled LCB molecule"
    ),
    value,
    contributing_species = species
  ) |>
  mutate(
    pseudovalue = pseudovalue,
    value_log10 = log10(value + pseudovalue),
    species = factor(species, levels = lcb_species_levels),
    lipid_class_short = factor(lipid_class_short, levels = "LCB")
  ) |>
  arrange(experiment, readout, species, replicate, time_h, sample_group)

flux_species_readouts <- flux_absolute_long |>
  filter(
    lipid_class_short %in% non_lcb_lipid_classes,
    label_group %in% c("labelled", "unlabelled"),
    !is_class_total,
    !is_global_total
  ) |>
  transmute(
    experiment,
    tracer,
    berlin_sample_id,
    sample_name,
    sample_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h,
    replicate,
    unit,
    lipid_class_short,
    species,
    species_base,
    label_status,
    label_group,
    readout = ifelse(
      label_group == "labelled",
      "labelled_incorporation",
      "unlabelled_abundance"
    ),
    readout_label = case_when(
      label_group == "labelled" & experiment == "Pal-d3" ~ "labelled_species_d3_plus_d6",
      label_group == "labelled" & experiment == "d7-dhSph" ~ "labelled_species_d7",
      TRUE ~ "Unlabelled lipid species"
    ),
    value,
    contributing_species = species
  ) |>
  mutate(
    pseudovalue = pseudovalue,
    value_log10 = log10(value + pseudovalue),
    lipid_class_short = factor(lipid_class_short, levels = non_lcb_lipid_classes)
  ) |>
  arrange(
    experiment,
    readout,
    lipid_class_short,
    species_sort_index(species),
    species,
    replicate,
    time_h,
    sample_group
  )

flux_analysis_readouts_unfactored <- bind_rows(
  flux_lcb_readouts |>
    mutate(
      feature_type = "LCB molecule",
      feature_name = format_analysis_feature(lipid_class_short, species)
    ),
  flux_species_readouts |>
    mutate(
      feature_type = "Lipid species",
      feature_name = format_analysis_feature(lipid_class_short, species)
    )
)

analysis_feature_levels <- flux_analysis_readouts_unfactored |>
  distinct(feature_type, lipid_class_short, species, feature_name) |>
  mutate(
    feature_type = factor(feature_type, levels = analysis_feature_type_levels),
    lipid_class_short = factor(
      lipid_class_short,
      levels = c("LCB", non_lcb_lipid_classes)
    ),
    species_order = species_sort_index(species)
  ) |>
  arrange(feature_type, lipid_class_short, species_order, feature_name) |>
  pull(feature_name) |>
  unique()

flux_analysis_readouts <- flux_analysis_readouts_unfactored |>
  mutate(
    feature_type = factor(feature_type, levels = analysis_feature_type_levels),
    feature_name = factor(feature_name, levels = analysis_feature_levels),
    readout = factor(
      readout,
      levels = c("labelled_incorporation", "unlabelled_abundance")
    )
  ) |>
  arrange(experiment, readout, feature_name, replicate, time_h, sample_group)

flux_labelled_species <- flux_absolute_long |>
  filter(
    label_group == "labelled",
    !is_class_total,
    !is_global_total
  ) |>
  mutate(
    value_log10 = log10(value + pseudovalue),
    lipid_class_short = factor(
      lipid_class_short,
      levels = c("LCB", "dhCer", "Cer", "dhSM", "SM", "HexCer", "LacCer")
    )
  ) |>
  arrange(experiment, lipid_class_short, species, replicate, time_h, sample_group)

labelled_fraction <- labelled_class |>
  select(
    experiment,
    tracer,
    berlin_sample_id,
    sample_name,
    sample_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h,
    replicate,
    lipid_class_short,
    labelled_value = value
  ) |>
  inner_join(
    class_total |>
      select(
        experiment,
        berlin_sample_id,
        lipid_class_short,
        class_total_value = value
      ),
    by = c("experiment", "berlin_sample_id", "lipid_class_short")
  ) |>
  mutate(
    labelled_fraction = ifelse(
      class_total_value > 0,
      labelled_value / class_total_value,
      NA_real_
    )
  )

write_csv_if_requested(sample_design, "sample_design.csv")
write_csv_if_requested(flux_long, "flux_long.csv")
write_csv_if_requested(flux_absolute_long, "flux_absolute_long.csv")
write_csv_if_requested(flux_class_readouts, "flux_class_readouts.csv")
write_csv_if_requested(flux_lcb_readouts, "flux_lcb_readouts.csv")
write_csv_if_requested(flux_species_readouts, "flux_species_readouts.csv")
write_csv_if_requested(flux_analysis_readouts, "flux_analysis_readouts.csv")
write_csv_if_requested(flux_labelled_species, "flux_labelled_species.csv")
write_csv_if_requested(labelled_fraction, "labelled_fraction.csv")

list(
  sample_design_rows = nrow(sample_design),
  flux_long_rows = nrow(flux_long),
  flux_absolute_long_rows = nrow(flux_absolute_long),
  flux_class_readout_rows = nrow(flux_class_readouts),
  flux_lcb_readout_rows = nrow(flux_lcb_readouts),
  flux_species_readout_rows = nrow(flux_species_readouts),
  flux_analysis_readout_rows = nrow(flux_analysis_readouts),
  labelled_species_rows = nrow(flux_labelled_species)
)
```

## 6. Import Checks

```{r import-checks}
sample_design |>
  count(experiment, tracer_treatment, infection, time_h) |>
  arrange(experiment, tracer_treatment, infection, time_h)

flux_absolute_long |>
  count(experiment, lipid_class_short, label_status) |>
  arrange(experiment, lipid_class_short, label_status)

no_tracer_label_check <- flux_labelled_species |>
  filter(!is_tracer_treated) |>
  group_by(experiment) |>
  summarise(
    nonzero_labelled_rows = sum(!is.na(value) & value != 0),
    max_labelled_value = max(value, na.rm = TRUE),
    .groups = "drop"
  )

no_tracer_label_check
```

## 7. Time-Course Summaries

```{r timecourse-summaries}
mean_se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

flux_class_summary <- flux_class_readouts |>
  group_by(
    experiment,
    tracer,
    readout,
    readout_label,
    lipid_class_short,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h
  ) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    se_value = mean_se(value),
    ci_95 = qt(0.975, df = pmax(n - 1, 1)) * se_value,
    ci_low = pmax(mean_value - ci_95, 0),
    ci_high = mean_value + ci_95,
    mean_log10 = mean(value_log10, na.rm = TRUE),
    se_log10 = mean_se(value_log10),
    .groups = "drop"
  )

flux_lcb_summary <- flux_lcb_readouts |>
  group_by(
    experiment,
    tracer,
    readout,
    readout_label,
    lipid_class_short,
    species,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h
  ) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    se_value = mean_se(value),
    ci_95 = qt(0.975, df = pmax(n - 1, 1)) * se_value,
    ci_low = pmax(mean_value - ci_95, 0),
    ci_high = mean_value + ci_95,
    mean_log10 = mean(value_log10, na.rm = TRUE),
    se_log10 = mean_se(value_log10),
    .groups = "drop"
  )

flux_species_summary <- flux_species_readouts |>
  group_by(
    experiment,
    tracer,
    readout,
    readout_label,
    lipid_class_short,
    species,
    species_base,
    label_status,
    label_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h
  ) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    se_value = mean_se(value),
    ci_95 = qt(0.975, df = pmax(n - 1, 1)) * se_value,
    ci_low = pmax(mean_value - ci_95, 0),
    ci_high = mean_value + ci_95,
    mean_log10 = mean(value_log10, na.rm = TRUE),
    se_log10 = mean_se(value_log10),
    .groups = "drop"
  )

flux_analysis_summary <- flux_analysis_readouts |>
  group_by(
    experiment,
    tracer,
    readout,
    readout_label,
    feature_type,
    feature_name,
    lipid_class_short,
    species,
    species_base,
    label_status,
    label_group,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h
  ) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    se_value = mean_se(value),
    ci_95 = qt(0.975, df = pmax(n - 1, 1)) * se_value,
    ci_low = pmax(mean_value - ci_95, 0),
    ci_high = mean_value + ci_95,
    mean_log10 = mean(value_log10, na.rm = TRUE),
    se_log10 = mean_se(value_log10),
    contributing_species = paste(sort(unique(contributing_species)), collapse = ";"),
    .groups = "drop"
  )

flux_labelled_species_summary <- flux_labelled_species |>
  group_by(
    experiment,
    tracer,
    lipid_class_short,
    species,
    infection,
    tracer_treatment,
    is_tracer_treated,
    time_h
  ) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    se_value = mean_se(value),
    ci_95 = qt(0.975, df = pmax(n - 1, 1)) * se_value,
    ci_low = pmax(mean_value - ci_95, 0),
    ci_high = mean_value + ci_95,
    mean_log10 = mean(value_log10, na.rm = TRUE),
    se_log10 = mean_se(value_log10),
    .groups = "drop"
  )

write_csv_if_requested(flux_class_summary, "flux_class_summary.csv")
write_csv_if_requested(flux_lcb_summary, "flux_lcb_summary.csv")
write_csv_if_requested(flux_species_summary, "flux_species_summary.csv")
write_csv_if_requested(flux_analysis_summary, "flux_analysis_summary.csv")
write_csv_if_requested(flux_labelled_species_summary, "flux_labelled_species_summary.csv")

flux_analysis_summary |>
  filter(readout == "labelled_incorporation", is_tracer_treated) |>
  select(
    experiment,
    feature_type,
    feature_name,
    infection,
    time_h,
    mean_value,
    se_value,
    n
  ) |>
  arrange(experiment, feature_name, time_h, infection)
```

## 8. AUC Analysis

AUC is calculated with the trapezoid rule from the observed timepoints
0.5-8 h. This summarizes total labelled incorporation across the time course.
It should be interpreted alongside the time-course plots, not as a replacement
for the time element.

```{r auc-analysis}
trapezoid_auc <- function(time, value) {
  ok <- !is.na(time) & !is.na(value)
  time <- time[ok]
  value <- value[ok]

  if (length(time) < 2) {
    return(NA_real_)
  }

  order_index <- order(time)
  time <- time[order_index]
  value <- value[order_index]

  sum(diff(time) * (head(value, -1) + tail(value, -1)) / 2)
}

auc_by_replicate <- flux_analysis_readouts |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  group_by(
    experiment,
    tracer,
    feature_type,
    feature_name,
    infection,
    replicate
  ) |>
  summarise(
    auc_value_h = trapezoid_auc(time_h, value),
    auc_log10_h = trapezoid_auc(time_h, value_log10),
    .groups = "drop"
  )

auc_summary <- auc_by_replicate |>
  group_by(experiment, tracer, feature_type, feature_name, infection) |>
  summarise(
    mean_auc = mean(auc_value_h, na.rm = TRUE),
    sd_auc = sd(auc_value_h, na.rm = TRUE),
    se_auc = mean_se(auc_value_h),
    n = sum(!is.na(auc_value_h)),
    .groups = "drop"
  )

auc_ratio_summary <- auc_summary |>
  select(experiment, tracer, feature_type, feature_name, infection, mean_auc, se_auc, n) |>
  pivot_wider(
    names_from = infection,
    values_from = c(mean_auc, se_auc, n)
  ) |>
  mutate(
    infected_over_uninfected_auc = mean_auc_Infected / mean_auc_Uninfected
  ) |>
  arrange(experiment, feature_name)

write_csv_if_requested(auc_by_replicate, "auc_by_replicate.csv")
write_csv_if_requested(auc_summary, "auc_summary.csv")
write_csv_if_requested(auc_ratio_summary, "auc_ratio_summary.csv")

auc_ratio_summary
```

## 9. Timepoint-Level Tests

These first-pass tests compare infected versus uninfected samples at each
timepoint within each tracer and analysis feature. They fit a linear model on
`log10(value + pseudovalue)`, then use `emmeans` to extract the explicit
pairwise contrast `Infected - Uninfected`. The raw pairwise p-values are
calculated once, then two separate FDR chunks apply different multiple-testing
families. The primary table uses experiment-level BH-FDR across all
species/timepoint tests within each tracer experiment because these lipid
features are biologically related. A panel-level correction is retained only as
a secondary sensitivity view.

```{r timepoint-pairwise-tests}
empty_timepoint_pairwise <- function() {
  tibble(
    contrast = "Infected - Uninfected",
    estimate_log10_infected_minus_uninfected = NA_real_,
    se_log10 = NA_real_,
    df = NA_real_,
    t_ratio = NA_real_,
    lower_cl_log10 = NA_real_,
    upper_cl_log10 = NA_real_,
    ratio_infected_over_uninfected = NA_real_,
    ratio_lower_95 = NA_real_,
    ratio_upper_95 = NA_real_,
    p_value = NA_real_,
    pairwise_method = "emmeans pairwise contrast",
    model_status = "not_estimable"
  )
}

safe_infection_pairwise <- function(data) {
  data <- data |>
    filter(!is.na(value_log10), !is.na(infection)) |>
    mutate(infection = factor(infection, levels = c("Uninfected", "Infected")))

  if (n_distinct(data$infection) < 2 || nrow(data) < 4) {
    return(empty_timepoint_pairwise())
  }

  fit <- tryCatch(
    lm(value_log10 ~ infection, data = data),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(empty_timepoint_pairwise())
  }

  residual_deviance <- deviance(fit)
  response_scale <- max(1, sum(data$value_log10^2, na.rm = TRUE))
  if (residual_deviance <= 100 * .Machine$double.eps * response_scale) {
    estimate <- mean(data$value_log10[data$infection == "Infected"]) -
      mean(data$value_log10[data$infection == "Uninfected"])
    return(tibble(
      contrast = "Infected - Uninfected",
      estimate_log10_infected_minus_uninfected = estimate,
      se_log10 = NA_real_,
      df = NA_real_,
      t_ratio = NA_real_,
      lower_cl_log10 = NA_real_,
      upper_cl_log10 = NA_real_,
      ratio_infected_over_uninfected = 10^estimate,
      ratio_lower_95 = NA_real_,
      ratio_upper_95 = NA_real_,
      p_value = NA_real_,
      pairwise_method = "not estimated: zero or near-zero residual variance",
      model_status = "zero_or_near_zero_residual_variance"
    ))
  }

  em <- tryCatch(
    emmeans(fit, ~ infection),
    error = function(e) NULL
  )

  if (is.null(em)) {
    return(empty_timepoint_pairwise())
  }

  pairwise <- tryCatch(
    contrast(
      em,
      method = list("Infected - Uninfected" = c(-1, 1)),
      adjust = "none"
    ),
    error = function(e) NULL
  )

  if (is.null(pairwise)) {
    return(empty_timepoint_pairwise())
  }

  pairwise_summary <- summary(pairwise, infer = c(TRUE, TRUE)) |>
    as_tibble()

  if (nrow(pairwise_summary) == 0) {
    return(empty_timepoint_pairwise())
  }

  get_pairwise_col <- function(name) {
    if (name %in% names(pairwise_summary)) {
      pairwise_summary[[name]][1]
    } else {
      NA_real_
    }
  }

  estimate <- get_pairwise_col("estimate")
  lower_cl <- get_pairwise_col("lower.CL")
  upper_cl <- get_pairwise_col("upper.CL")

  tibble(
    contrast = as.character(pairwise_summary$contrast[1]),
    estimate_log10_infected_minus_uninfected = estimate,
    se_log10 = get_pairwise_col("SE"),
    df = get_pairwise_col("df"),
    t_ratio = get_pairwise_col("t.ratio"),
    lower_cl_log10 = lower_cl,
    upper_cl_log10 = upper_cl,
    ratio_infected_over_uninfected = 10^estimate,
    ratio_lower_95 = 10^lower_cl,
    ratio_upper_95 = 10^upper_cl,
    p_value = get_pairwise_col("p.value"),
    pairwise_method = "emmeans pairwise contrast",
    model_status = "estimated"
  )
}

timepoint_pairwise_raw <- flux_analysis_readouts |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  group_by(experiment, tracer, feature_type, feature_name, time_h) |>
  group_modify(~ safe_infection_pairwise(.x)) |>
  ungroup() |>
  arrange(experiment, feature_name, time_h)

write_csv_if_requested(
  timepoint_pairwise_raw,
  "timepoint_pairwise_emmeans_raw_labelled_incorporation.csv"
)

timepoint_pairwise_raw
```

### 9.1 Experiment-Level Timepoint FDR

This is the primary timepoint table. All infected-versus-uninfected timepoint
comparisons within the same tracer experiment are treated as one
multiple-testing family.

For the selected Pal-d3 fractional-incorporation endpoints, a separate
absolute-abundance analysis pools the d3 and d6 16:0 species before these
timepoint tests. Run `scripts/pal_d3_pooled_absolute_analysis.R` after
generating `flux_absolute_long.csv`; see its `ANALYSIS_AND_RESULTS.md` output.
For pooled d7-labelled class endpoints, run
`scripts/d7_pooled_absolute_analysis.R`. The table below remains the
individual-species analysis.

```{r timepoint-experiment-fdr}
timepoint_tests <- timepoint_pairwise_raw |>
  group_by(experiment) |>
  mutate(
    p_fdr = p.adjust(p_value, method = "BH"),
    fdr_method = "BH within experiment across species and timepoints"
  ) |>
  ungroup() |>
  arrange(experiment, feature_name, time_h)

write_csv_if_requested(timepoint_tests, "timepoint_tests_labelled_incorporation.csv")
write_csv_if_requested(
  timepoint_tests,
  "timepoint_pairwise_emmeans_labelled_incorporation.csv"
)
write_csv_if_requested(
  timepoint_tests,
  "timepoint_tests_experiment_fdr_labelled_incorporation.csv"
)
write_csv_if_requested(
  timepoint_tests,
  "timepoint_pairwise_emmeans_experiment_fdr_labelled_incorporation.csv"
)

timepoint_tests
```

### 9.2 Panel-Level Timepoint FDR

This secondary table treats each plotted species/readout panel as its own
hypothesis family and applies BH-FDR across sampled timepoints within that
panel. It is kept for comparison with older panel-wise summaries.

```{r timepoint-panel-fdr}
timepoint_tests_panel_fdr <- timepoint_pairwise_raw |>
  group_by(experiment, tracer, feature_type, feature_name) |>
  mutate(
    p_fdr = p.adjust(p_value, method = "BH"),
    fdr_method = "BH within species panel across timepoints"
  ) |>
  ungroup() |>
  arrange(experiment, feature_name, time_h)

write_csv_if_requested(
  timepoint_tests_panel_fdr,
  "timepoint_tests_panel_fdr_labelled_incorporation.csv"
)
write_csv_if_requested(
  timepoint_tests_panel_fdr,
  "timepoint_pairwise_emmeans_panel_fdr_labelled_incorporation.csv"
)

timepoint_tests_panel_fdr
```

## 10. Infection-by-Time Models

These models ask whether each plotted readout differs by infection, time, and
infection-by-time interaction for each tracer and analysis feature. The plotted
`curve FDR` annotation comes from the `infection:time_f` term.

```{r infection-time-models}
safe_infection_time_model <- function(data) {
  data <- data |>
    filter(!is.na(value_log10), !is.na(infection), !is.na(time_h)) |>
    mutate(
      infection = factor(infection, levels = c("Uninfected", "Infected")),
      time_f = factor(time_h)
    )

  if (n_distinct(data$infection) < 2 || n_distinct(data$time_f) < 2) {
    return(tibble(
      p_infection = NA_real_,
      p_time = NA_real_,
      p_infection_time = NA_real_,
      model_status = "not_estimable"
    ))
  }

  fit <- tryCatch(
    lm(value_log10 ~ infection * time_f, data = data),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(tibble(
      p_infection = NA_real_,
      p_time = NA_real_,
      p_infection_time = NA_real_,
      model_status = "not_estimable"
    ))
  }

  residual_deviance <- deviance(fit)
  response_scale <- max(1, sum(data$value_log10^2, na.rm = TRUE))
  if (residual_deviance <= 100 * .Machine$double.eps * response_scale) {
    return(tibble(
      p_infection = NA_real_,
      p_time = NA_real_,
      p_infection_time = NA_real_,
      model_status = "zero_or_near_zero_residual_variance"
    ))
  }

  a <- anova(fit)

  tibble(
    p_infection = a["infection", "Pr(>F)"],
    p_time = a["time_f", "Pr(>F)"],
    p_infection_time = a["infection:time_f", "Pr(>F)"],
    model_status = "estimated"
  )
}

infection_time_models_all_readouts <- flux_analysis_readouts |>
  filter(
    readout %in% c("labelled_incorporation", "unlabelled_abundance"),
    is_tracer_treated
  ) |>
  group_by(experiment, tracer, readout, readout_label, feature_type, feature_name) |>
  group_modify(~ safe_infection_time_model(.x)) |>
  ungroup() |>
  group_by(experiment, readout) |>
  mutate(
    p_fdr_infection = p.adjust(p_infection, method = "BH"),
    p_fdr_time = p.adjust(p_time, method = "BH"),
    p_fdr_infection_time = p.adjust(p_infection_time, method = "BH")
  ) |>
  ungroup() |>
  arrange(experiment, readout, feature_name)

curve_fdr_labels <- infection_time_models_all_readouts |>
  transmute(
    experiment,
    tracer,
    readout,
    feature_type,
    feature_name,
    curve_fdr = p_fdr_infection_time,
    curve_fdr_label = format_curve_fdr_label(p_fdr_infection_time)
  )

infection_time_models <- infection_time_models_all_readouts |>
  filter(readout == "labelled_incorporation") |>
  select(
    experiment,
    tracer,
    feature_type,
    feature_name,
    p_infection,
    p_time,
    p_infection_time,
    p_fdr_infection,
    p_fdr_time,
    p_fdr_infection_time
  )

write_csv_if_requested(
  infection_time_models_all_readouts,
  "infection_time_models_all_readouts.csv"
)
write_csv_if_requested(infection_time_models, "infection_time_models.csv")
write_csv_if_requested(curve_fdr_labels, "curve_fdr_labels.csv")

infection_time_models
```

## 10.1 Normalized Block Sensitivity Analysis

This section repeats the main labelled/unlabelled workflow on the normalized
`fmol / pmol total sphingolipids` block when that block is imported. It is a
sensitivity check only: normalized results are kept in separate output tables
and are compared back to the primary absolute `pmol / sample` results.

```{r normalized-sensitivity-analysis}
make_analysis_readouts_from_long <- function(source_long) {
  empty <- list(
    class_readouts = tibble(),
    lcb_readouts = tibble(),
    species_readouts = tibble(),
    analysis_readouts = tibble(),
    labelled_species = tibble()
  )

  if (nrow(source_long) == 0) {
    return(empty)
  }

  unit_unlabelled_class <- source_long |>
    filter(
      lipid_class_short %in% non_lcb_lipid_classes,
      label_group == "unlabelled",
      !is_class_total,
      !is_global_total
    ) |>
    group_by(
      experiment,
      tracer,
      berlin_sample_id,
      sample_name,
      sample_group,
      infection,
      tracer_treatment,
      is_tracer_treated,
      time_h,
      replicate,
      unit,
      lipid_class_short
    ) |>
    summarise(
      readout = "unlabelled_abundance",
      readout_label = "Unlabelled abundance",
      value = sum(value, na.rm = TRUE),
      contributing_species = paste(sort(unique(species)), collapse = ";"),
      .groups = "drop"
    )

  unit_labelled_class <- source_long |>
    filter(
      lipid_class_short %in% non_lcb_lipid_classes,
      label_group == "labelled",
      !is_class_total,
      !is_global_total
    ) |>
    group_by(
      experiment,
      tracer,
      berlin_sample_id,
      sample_name,
      sample_group,
      infection,
      tracer_treatment,
      is_tracer_treated,
      time_h,
      replicate,
      unit,
      lipid_class_short
    ) |>
    summarise(
      readout = "labelled_incorporation",
      readout_label = ifelse(
        first(experiment) == "Pal-d3",
        "labelled_flux_d3_plus_d6",
        "labelled_flux_d7"
      ),
      value = sum(value, na.rm = TRUE),
      contributing_species = paste(sort(unique(species)), collapse = ";"),
      .groups = "drop"
    )

  unit_class_total <- source_long |>
    filter(
      lipid_class_short %in% non_lcb_lipid_classes,
      is_class_total,
      !is_global_total
    ) |>
    transmute(
      experiment,
      tracer,
      berlin_sample_id,
      sample_name,
      sample_group,
      infection,
      tracer_treatment,
      is_tracer_treated,
      time_h,
      replicate,
      unit,
      lipid_class_short,
      readout = "class_total_in_workbook",
      readout_label = "Class total in workbook",
      value,
      contributing_species = species
    )

  unit_class_readouts <- bind_rows(
    unit_unlabelled_class,
    unit_labelled_class,
    unit_class_total
  ) |>
    mutate(
      pseudovalue = pseudovalue,
      value_log10 = log10(value + pseudovalue),
      readout = factor(
        readout,
        levels = c(
          "labelled_incorporation",
          "unlabelled_abundance",
          "class_total_in_workbook"
        )
      ),
      lipid_class_short = factor(lipid_class_short, levels = non_lcb_lipid_classes)
    ) |>
    arrange(experiment, readout, lipid_class_short, replicate, time_h, sample_group)

  unit_lcb_readouts <- source_long |>
    filter(
      lipid_class_short == "LCB",
      label_group %in% c("labelled", "unlabelled"),
      !is_class_total,
      !is_global_total
    ) |>
    transmute(
      experiment,
      tracer,
      berlin_sample_id,
      sample_name,
      sample_group,
      infection,
      tracer_treatment,
      is_tracer_treated,
      time_h,
      replicate,
      unit,
      lipid_class_short,
      species,
      species_base,
      label_status,
      label_group,
      readout = ifelse(
        label_group == "labelled",
        "labelled_incorporation",
        "unlabelled_abundance"
      ),
      readout_label = case_when(
        label_group == "labelled" & experiment == "Pal-d3" ~ "labelled_LCB_d3",
        label_group == "labelled" & experiment == "d7-dhSph" ~ "labelled_LCB_d7",
        TRUE ~ "Unlabelled LCB molecule"
      ),
      value,
      contributing_species = species
    ) |>
    mutate(
      pseudovalue = pseudovalue,
      value_log10 = log10(value + pseudovalue),
      species = factor(species, levels = lcb_species_levels),
      lipid_class_short = factor(lipid_class_short, levels = "LCB")
    ) |>
    arrange(experiment, readout, species, replicate, time_h, sample_group)

  unit_species_readouts <- source_long |>
    filter(
      lipid_class_short %in% non_lcb_lipid_classes,
      label_group %in% c("labelled", "unlabelled"),
      !is_class_total,
      !is_global_total
    ) |>
    transmute(
      experiment,
      tracer,
      berlin_sample_id,
      sample_name,
      sample_group,
      infection,
      tracer_treatment,
      is_tracer_treated,
      time_h,
      replicate,
      unit,
      lipid_class_short,
      species,
      species_base,
      label_status,
      label_group,
      readout = ifelse(
        label_group == "labelled",
        "labelled_incorporation",
        "unlabelled_abundance"
      ),
      readout_label = case_when(
        label_group == "labelled" & experiment == "Pal-d3" ~ "labelled_species_d3_plus_d6",
        label_group == "labelled" & experiment == "d7-dhSph" ~ "labelled_species_d7",
        TRUE ~ "Unlabelled lipid species"
      ),
      value,
      contributing_species = species
    ) |>
    mutate(
      pseudovalue = pseudovalue,
      value_log10 = log10(value + pseudovalue),
      lipid_class_short = factor(lipid_class_short, levels = non_lcb_lipid_classes)
    ) |>
    arrange(
      experiment,
      readout,
      lipid_class_short,
      species_sort_index(species),
      species,
      replicate,
      time_h,
      sample_group
    )

  unit_analysis_readouts_unfactored <- bind_rows(
    unit_lcb_readouts |>
      mutate(
        feature_type = "LCB molecule",
        feature_name = format_analysis_feature(lipid_class_short, species)
      ),
    unit_species_readouts |>
      mutate(
        feature_type = "Lipid species",
        feature_name = format_analysis_feature(lipid_class_short, species)
      )
  )

  unit_analysis_feature_levels <- unit_analysis_readouts_unfactored |>
    distinct(feature_type, lipid_class_short, species, feature_name) |>
    mutate(
      feature_type = factor(feature_type, levels = analysis_feature_type_levels),
      lipid_class_short = factor(
        lipid_class_short,
        levels = c("LCB", non_lcb_lipid_classes)
      ),
      species_order = species_sort_index(species)
    ) |>
    arrange(feature_type, lipid_class_short, species_order, feature_name) |>
    pull(feature_name) |>
    unique()

  unit_analysis_readouts <- unit_analysis_readouts_unfactored |>
    mutate(
      feature_type = factor(feature_type, levels = analysis_feature_type_levels),
      feature_name = factor(feature_name, levels = unit_analysis_feature_levels),
      readout = factor(
        readout,
        levels = c("labelled_incorporation", "unlabelled_abundance")
      )
    ) |>
    arrange(experiment, readout, feature_name, replicate, time_h, sample_group)

  unit_labelled_species <- source_long |>
    filter(
      label_group == "labelled",
      !is_class_total,
      !is_global_total
    ) |>
    mutate(
      value_log10 = log10(value + pseudovalue),
      lipid_class_short = factor(
        lipid_class_short,
        levels = c("LCB", "dhCer", "Cer", "dhSM", "SM", "HexCer", "LacCer")
      )
    ) |>
    arrange(experiment, lipid_class_short, species, replicate, time_h, sample_group)

  list(
    class_readouts = unit_class_readouts,
    lcb_readouts = unit_lcb_readouts,
    species_readouts = unit_species_readouts,
    analysis_readouts = unit_analysis_readouts,
    labelled_species = unit_labelled_species
  )
}

classify_ratio_direction <- function(x) {
  case_when(
    is.na(x) ~ NA_character_,
    x > 1 ~ "infected_higher",
    x < 1 ~ "infected_lower",
    TRUE ~ "equal"
  )
}

agreement_summary <- function(data, agreement_col, label) {
  data |>
    summarise(
      comparison = label,
      n_compared = sum(!is.na(.data[[agreement_col]])),
      n_agree = sum(.data[[agreement_col]], na.rm = TRUE),
      pct_agree = ifelse(n_compared > 0, 100 * n_agree / n_compared, NA_real_)
    )
}

flux_normalized_long <- flux_long |>
  filter(unit == "fmol_per_pmol_total_sphingolipids")

normalized_tables <- make_analysis_readouts_from_long(flux_normalized_long)
normalized_class_readouts <- normalized_tables$class_readouts
normalized_lcb_readouts <- normalized_tables$lcb_readouts
normalized_species_readouts <- normalized_tables$species_readouts
normalized_analysis_readouts <- normalized_tables$analysis_readouts
normalized_labelled_species <- normalized_tables$labelled_species

if (nrow(normalized_analysis_readouts) > 0) {
  normalized_auc_by_replicate <- normalized_analysis_readouts |>
    filter(
      readout == "labelled_incorporation",
      is_tracer_treated
    ) |>
    group_by(
      experiment,
      tracer,
      feature_type,
      feature_name,
      infection,
      replicate
    ) |>
    summarise(
      auc_value_h = trapezoid_auc(time_h, value),
      auc_log10_h = trapezoid_auc(time_h, value_log10),
      .groups = "drop"
    )

  normalized_auc_summary <- normalized_auc_by_replicate |>
    group_by(experiment, tracer, feature_type, feature_name, infection) |>
    summarise(
      mean_auc = mean(auc_value_h, na.rm = TRUE),
      sd_auc = sd(auc_value_h, na.rm = TRUE),
      se_auc = mean_se(auc_value_h),
      n = sum(!is.na(auc_value_h)),
      .groups = "drop"
    )

  normalized_auc_ratio_summary <- normalized_auc_summary |>
    select(experiment, tracer, feature_type, feature_name, infection, mean_auc, se_auc, n) |>
    pivot_wider(
      names_from = infection,
      values_from = c(mean_auc, se_auc, n)
    ) |>
    mutate(
      infected_over_uninfected_auc = mean_auc_Infected / mean_auc_Uninfected
    ) |>
    arrange(experiment, feature_name)

  normalized_timepoint_pairwise_raw <- normalized_analysis_readouts |>
    filter(
      readout == "labelled_incorporation",
      is_tracer_treated
    ) |>
    group_by(experiment, tracer, feature_type, feature_name, time_h) |>
    group_modify(~ safe_infection_pairwise(.x)) |>
    ungroup() |>
    arrange(experiment, feature_name, time_h)

  normalized_timepoint_tests <- normalized_timepoint_pairwise_raw |>
    group_by(experiment) |>
    mutate(
      p_fdr = p.adjust(p_value, method = "BH"),
      fdr_method = "BH within experiment across species and timepoints"
    ) |>
    ungroup() |>
    arrange(experiment, feature_name, time_h)

  normalized_timepoint_tests_panel_fdr <- normalized_timepoint_pairwise_raw |>
    group_by(experiment, tracer, feature_type, feature_name) |>
    mutate(
      p_fdr = p.adjust(p_value, method = "BH"),
      fdr_method = "BH within species panel across timepoints"
    ) |>
    ungroup() |>
    arrange(experiment, feature_name, time_h)

  normalized_infection_time_models_all_readouts <- normalized_analysis_readouts |>
    filter(
      readout %in% c("labelled_incorporation", "unlabelled_abundance"),
      is_tracer_treated
    ) |>
    group_by(experiment, tracer, readout, readout_label, feature_type, feature_name) |>
    group_modify(~ safe_infection_time_model(.x)) |>
    ungroup() |>
    group_by(experiment, readout) |>
    mutate(
      p_fdr_infection = p.adjust(p_infection, method = "BH"),
      p_fdr_time = p.adjust(p_time, method = "BH"),
      p_fdr_infection_time = p.adjust(p_infection_time, method = "BH")
    ) |>
    ungroup() |>
    arrange(experiment, readout, feature_name)

  normalized_curve_fdr_labels <- normalized_infection_time_models_all_readouts |>
    transmute(
      experiment,
      tracer,
      readout,
      feature_type,
      feature_name,
      normalized_curve_fdr = p_fdr_infection_time,
      normalized_curve_fdr_label = format_curve_fdr_label(p_fdr_infection_time)
    )

  curve_fdr_sensitivity_comparison <- curve_fdr_labels |>
    rename(
      absolute_curve_fdr = curve_fdr,
      absolute_curve_fdr_label = curve_fdr_label
    ) |>
    full_join(
      normalized_curve_fdr_labels,
      by = c("experiment", "tracer", "readout", "feature_type", "feature_name")
    ) |>
    mutate(
      absolute_significant = absolute_curve_fdr < 0.05,
      normalized_significant = normalized_curve_fdr < 0.05,
      same_significance_call = ifelse(
        is.na(absolute_significant) | is.na(normalized_significant),
        NA,
        absolute_significant == normalized_significant
      )
    ) |>
    arrange(experiment, readout, feature_name)

  auc_sensitivity_comparison <- auc_ratio_summary |>
    transmute(
      experiment,
      tracer,
      feature_type,
      feature_name,
      absolute_auc_ratio = infected_over_uninfected_auc,
      absolute_auc_direction = classify_ratio_direction(infected_over_uninfected_auc)
    ) |>
    full_join(
      normalized_auc_ratio_summary |>
        transmute(
          experiment,
          tracer,
          feature_type,
          feature_name,
          normalized_auc_ratio = infected_over_uninfected_auc,
          normalized_auc_direction = classify_ratio_direction(infected_over_uninfected_auc)
        ),
      by = c("experiment", "tracer", "feature_type", "feature_name")
    ) |>
    mutate(
      same_auc_direction = ifelse(
        is.na(absolute_auc_direction) | is.na(normalized_auc_direction),
        NA,
        absolute_auc_direction == normalized_auc_direction
      )
    ) |>
    arrange(experiment, feature_name)

  timepoint_sensitivity_comparison <- timepoint_tests |>
    transmute(
      experiment,
      tracer,
      feature_type,
      feature_name,
      time_h,
      absolute_ratio = ratio_infected_over_uninfected,
      absolute_fdr = p_fdr,
      absolute_significant = p_fdr < 0.05
    ) |>
    full_join(
      normalized_timepoint_tests |>
        transmute(
          experiment,
          tracer,
          feature_type,
          feature_name,
          time_h,
          normalized_ratio = ratio_infected_over_uninfected,
          normalized_fdr = p_fdr,
          normalized_significant = p_fdr < 0.05
        ),
      by = c("experiment", "tracer", "feature_type", "feature_name", "time_h")
    ) |>
    mutate(
      same_significance_call = ifelse(
        is.na(absolute_significant) | is.na(normalized_significant),
        NA,
        absolute_significant == normalized_significant
      )
    ) |>
    arrange(experiment, feature_name, time_h)

  timepoint_sensitivity_comparison_panel_fdr <- timepoint_tests_panel_fdr |>
    transmute(
      experiment,
      tracer,
      feature_type,
      feature_name,
      time_h,
      absolute_ratio = ratio_infected_over_uninfected,
      absolute_fdr = p_fdr,
      absolute_significant = p_fdr < 0.05
    ) |>
    full_join(
      normalized_timepoint_tests_panel_fdr |>
        transmute(
          experiment,
          tracer,
          feature_type,
          feature_name,
          time_h,
          normalized_ratio = ratio_infected_over_uninfected,
          normalized_fdr = p_fdr,
          normalized_significant = p_fdr < 0.05
        ),
      by = c("experiment", "tracer", "feature_type", "feature_name", "time_h")
    ) |>
    mutate(
      same_significance_call = ifelse(
        is.na(absolute_significant) | is.na(normalized_significant),
        NA,
        absolute_significant == normalized_significant
      )
    ) |>
    arrange(experiment, feature_name, time_h)

  normalized_sensitivity_summary <- bind_rows(
    agreement_summary(
      curve_fdr_sensitivity_comparison,
      "same_significance_call",
      "Curve FDR significance agreement, labelled and unlabelled readouts"
    ),
    agreement_summary(
      curve_fdr_sensitivity_comparison |>
        filter(readout == "labelled_incorporation"),
      "same_significance_call",
      "Curve FDR significance agreement, labelled incorporation only"
    ),
    agreement_summary(
      timepoint_sensitivity_comparison,
      "same_significance_call",
      "Timepoint experiment-level FDR significance agreement, labelled incorporation"
    ),
    agreement_summary(
      timepoint_sensitivity_comparison_panel_fdr,
      "same_significance_call",
      "Timepoint panel-level FDR significance agreement, labelled incorporation"
    ),
    agreement_summary(
      auc_sensitivity_comparison,
      "same_auc_direction",
      "AUC infected/uninfected direction agreement, labelled incorporation"
    )
  )
} else {
  normalized_auc_by_replicate <- tibble()
  normalized_auc_summary <- tibble()
  normalized_auc_ratio_summary <- tibble()
  normalized_timepoint_pairwise_raw <- tibble()
  normalized_timepoint_tests <- tibble()
  normalized_timepoint_tests_panel_fdr <- tibble()
  normalized_infection_time_models_all_readouts <- tibble()
  normalized_curve_fdr_labels <- tibble()
  curve_fdr_sensitivity_comparison <- tibble()
  auc_sensitivity_comparison <- tibble()
  timepoint_sensitivity_comparison <- tibble()
  timepoint_sensitivity_comparison_panel_fdr <- tibble()
  normalized_sensitivity_summary <- tibble(
    comparison = "Normalized block was not imported",
    n_compared = 0,
    n_agree = 0,
    pct_agree = NA_real_
  )
}

write_csv_if_requested(flux_normalized_long, "normalized_sensitivity_flux_long.csv")
write_csv_if_requested(
  normalized_class_readouts,
  "normalized_sensitivity_class_readouts.csv"
)
write_csv_if_requested(
  normalized_lcb_readouts,
  "normalized_sensitivity_lcb_readouts.csv"
)
write_csv_if_requested(
  normalized_species_readouts,
  "normalized_sensitivity_species_readouts.csv"
)
write_csv_if_requested(
  normalized_analysis_readouts,
  "normalized_sensitivity_analysis_readouts.csv"
)
write_csv_if_requested(
  normalized_labelled_species,
  "normalized_sensitivity_labelled_species.csv"
)
write_csv_if_requested(
  normalized_auc_by_replicate,
  "normalized_sensitivity_auc_by_replicate.csv"
)
write_csv_if_requested(
  normalized_auc_summary,
  "normalized_sensitivity_auc_summary.csv"
)
write_csv_if_requested(
  normalized_auc_ratio_summary,
  "normalized_sensitivity_auc_ratio_summary.csv"
)
write_csv_if_requested(
  normalized_timepoint_pairwise_raw,
  "normalized_sensitivity_timepoint_pairwise_emmeans_raw_labelled_incorporation.csv"
)
write_csv_if_requested(
  normalized_timepoint_tests,
  "normalized_sensitivity_timepoint_pairwise_emmeans_labelled_incorporation.csv"
)
write_csv_if_requested(
  normalized_timepoint_tests,
  "normalized_sensitivity_timepoint_pairwise_emmeans_experiment_fdr_labelled_incorporation.csv"
)
write_csv_if_requested(
  normalized_timepoint_tests_panel_fdr,
  "normalized_sensitivity_timepoint_pairwise_emmeans_panel_fdr_labelled_incorporation.csv"
)
write_csv_if_requested(
  normalized_infection_time_models_all_readouts,
  "normalized_sensitivity_infection_time_models_all_readouts.csv"
)
write_csv_if_requested(
  normalized_curve_fdr_labels,
  "normalized_sensitivity_curve_fdr_labels.csv"
)
write_csv_if_requested(
  curve_fdr_sensitivity_comparison,
  "normalized_sensitivity_curve_fdr_comparison.csv"
)
write_csv_if_requested(
  auc_sensitivity_comparison,
  "normalized_sensitivity_auc_comparison.csv"
)
write_csv_if_requested(
  timepoint_sensitivity_comparison,
  "normalized_sensitivity_timepoint_comparison.csv"
)
write_csv_if_requested(
  timepoint_sensitivity_comparison_panel_fdr,
  "normalized_sensitivity_timepoint_comparison_panel_fdr.csv"
)
write_csv_if_requested(
  normalized_sensitivity_summary,
  "normalized_sensitivity_summary.csv"
)

if (write_outputs) {
  summary_file <- file.path(output_dir, "normalized_sensitivity_summary.md")
  writeLines(
    c(
      "# Normalized Block Sensitivity Analysis",
      "",
      "This sensitivity analysis repeats the main isotope-tracer statistics on the normalized `fmol / pmol total sphingolipids` block.",
      "The primary analysis remains the absolute `pmol / sample` block.",
      "",
      "Both analyses use the same model structure:",
      "",
      "```text",
      "log10(value + 0.01) ~ infection * time_f",
      "```",
      "",
      "The main comparison is whether the BH-adjusted infection-by-time significance call agrees between absolute and normalized measurements.",
      "",
      paste(capture.output(print(normalized_sensitivity_summary)), collapse = "\n")
    ),
    con = summary_file
  )
}

normalized_sensitivity_summary
```

## 11. Distribution Diagnostics

These diagnostics check whether the labelled incorporation features are strongly
right-skewed, whether low/zero values are common, and whether the current
log10-transformed infection-by-time models have reasonable residual behavior.
LCB entries are treated as molecule-level features, while non-LCB lipids are
treated as species-level features.

```{r distribution-diagnostic-data}
diagnostic_labelled <- flux_analysis_readouts |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  mutate(
    value_log10 = log10(value + pseudovalue),
    near_zero = value <= pseudovalue,
    experiment = factor(experiment, levels = c("Pal-d3", "d7-dhSph"))
  )

zero_nearzero_summary <- diagnostic_labelled |>
  group_by(experiment, feature_type, feature_name) |>
  summarise(
    n = n(),
    zero_n = sum(value == 0, na.rm = TRUE),
    near_zero_n = sum(near_zero, na.rm = TRUE),
    zero_pct = 100 * zero_n / n,
    near_zero_pct = 100 * near_zero_n / n,
    .groups = "drop"
  )

mean_sd_diagnostics <- diagnostic_labelled |>
  group_by(experiment, feature_type, feature_name, infection, time_h) |>
  summarise(
    mean_raw = mean(value, na.rm = TRUE),
    sd_raw = sd(value, na.rm = TRUE),
    mean_log10 = mean(value_log10, na.rm = TRUE),
    sd_log10 = sd(value_log10, na.rm = TRUE),
    n = sum(!is.na(value)),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(mean_raw, sd_raw, mean_log10, sd_log10),
    names_to = c(".value", "scale"),
    names_pattern = "(mean|sd)_(raw|log10)"
  ) |>
  mutate(
    scale = recode(scale, raw = "Raw pmol / sample", log10 = "log10(value + 0.01)")
  )

model_residual_diagnostics <- diagnostic_labelled |>
  group_by(experiment, feature_type, feature_name) |>
  group_modify(~ {
    model_data <- .x |>
      filter(!is.na(value_log10), !is.na(infection), !is.na(time_h)) |>
      mutate(
        infection = factor(infection, levels = c("Uninfected", "Infected")),
        time_f = factor(time_h)
      )

    fit <- lm(value_log10 ~ infection * time_f, data = model_data)

    tibble(
      infection = model_data$infection,
      time_h = model_data$time_h,
      fitted = fitted(fit),
      residual = resid(fit),
      std_residual = rstandard(fit)
    )
  }) |>
  ungroup()

write_csv_if_requested(zero_nearzero_summary, "diagnostics_zero_nearzero_summary.csv")
write_csv_if_requested(mean_sd_diagnostics, "diagnostics_mean_sd_summary.csv")
write_csv_if_requested(model_residual_diagnostics, "diagnostics_model_residuals.csv")

zero_nearzero_summary
```

### 11.1 Raw And Log10 Distributions

```{r plot-diagnostics-raw-log-distribution, fig.width=10, fig.height=5.5}
diagnostic_distribution_long <- diagnostic_labelled |>
  select(experiment, feature_type, feature_name, infection, value, value_log10) |>
  pivot_longer(
    cols = c(value, value_log10),
    names_to = "scale",
    values_to = "plot_value"
  ) |>
  mutate(
    scale = recode(
      scale,
      value = "Raw pmol / sample",
      value_log10 = "log10(value + 0.01)"
    )
  )

p_diagnostics_raw_log_distribution <- diagnostic_distribution_long |>
  ggplot(aes(x = plot_value, fill = infection, color = infection)) +
  geom_density(alpha = 0.18, linewidth = 0.5, na.rm = TRUE) +
  facet_grid(scale ~ experiment, scales = "free") +
  scale_color_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  scale_fill_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  labs(
    x = NULL,
    y = "Density",
    color = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    legend.position = "top"
  )

p_diagnostics_raw_log_distribution
save_plot_if_requested(
  p_diagnostics_raw_log_distribution,
  "diagnostics_raw_vs_log10_density.png",
  width = 10,
  height = 5.5
)
if (write_outputs) {
  ggsave(
    filename = file.path(output_dir, "diagnostics_raw_vs_log10_density.pdf"),
    plot = p_diagnostics_raw_log_distribution,
    width = 10,
    height = 5.5,
    units = "in",
    bg = "white"
  )
}
```

### 11.2 Zero And Near-Zero Burden

```{r plot-diagnostics-zero-nearzero, fig.width=8.5, fig.height=4.8}
p_diagnostics_zero_nearzero <- zero_nearzero_summary |>
  ggplot(aes(x = feature_name, y = near_zero_pct, fill = experiment)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "black") +
  labs(
    x = NULL,
    y = "Values <= pseudovalue (%)",
    fill = NULL
  ) +
  scale_fill_manual(values = c("Pal-d3" = "#56B4E9", "d7-dhSph" = "#E69F00")) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

p_diagnostics_zero_nearzero
save_plot_if_requested(
  p_diagnostics_zero_nearzero,
  "diagnostics_zero_nearzero_burden.png",
  width = 8.5,
  height = 4.8
)
if (write_outputs) {
  ggsave(
    filename = file.path(output_dir, "diagnostics_zero_nearzero_burden.pdf"),
    plot = p_diagnostics_zero_nearzero,
    width = 8.5,
    height = 4.8,
    units = "in",
    bg = "white"
  )
}
```

### 11.3 Mean-SD Relationship

```{r plot-diagnostics-mean-sd, fig.width=10, fig.height=5}
p_diagnostics_mean_sd <- mean_sd_diagnostics |>
  ggplot(aes(x = mean, y = sd, color = infection)) +
  geom_point(alpha = 0.75, size = 1.8) +
  facet_grid(scale ~ experiment, scales = "free") +
  scale_color_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  labs(
    x = "Mean",
    y = "SD",
    color = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    legend.position = "top"
  )

p_diagnostics_mean_sd
save_plot_if_requested(
  p_diagnostics_mean_sd,
  "diagnostics_mean_sd_raw_vs_log10.png",
  width = 10,
  height = 5
)
if (write_outputs) {
  ggsave(
    filename = file.path(output_dir, "diagnostics_mean_sd_raw_vs_log10.pdf"),
    plot = p_diagnostics_mean_sd,
    width = 10,
    height = 5,
    units = "in",
    bg = "white"
  )
}
```

### 11.4 Residuals Versus Fitted Values

```{r plot-diagnostics-residuals-fitted, fig.width=12, fig.height=7}
p_diagnostics_residuals_fitted <- model_residual_diagnostics |>
  ggplot(aes(x = fitted, y = residual, color = infection)) +
  geom_hline(yintercept = 0, linewidth = 0.35, color = "grey50") +
  geom_point(alpha = 0.75, size = 1.7) +
  facet_wrap(vars(experiment, feature_name), scales = "free", ncol = 6) +
  scale_color_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  labs(
    x = "Fitted log10(value + 0.01)",
    y = "Residual",
    color = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

p_diagnostics_residuals_fitted
save_plot_if_requested(
  p_diagnostics_residuals_fitted,
  "diagnostics_residuals_vs_fitted.png",
  width = 12,
  height = 7
)
if (write_outputs) {
  ggsave(
    filename = file.path(output_dir, "diagnostics_residuals_vs_fitted.pdf"),
    plot = p_diagnostics_residuals_fitted,
    width = 12,
    height = 7,
    units = "in",
    bg = "white"
  )
}
```

### 11.5 Q-Q Plot Of Model Residuals

```{r plot-diagnostics-qq, fig.width=12, fig.height=7}
p_diagnostics_qq <- model_residual_diagnostics |>
  ggplot(aes(sample = residual)) +
  stat_qq(alpha = 0.75, size = 1.7, color = "grey30") +
  stat_qq_line(linewidth = 0.35, color = "grey50") +
  facet_wrap(vars(experiment, feature_name), scales = "free", ncol = 6) +
  labs(
    x = "Theoretical normal quantile",
    y = "Sample residual quantile"
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    legend.position = "none"
  )

p_diagnostics_qq
save_plot_if_requested(
  p_diagnostics_qq,
  "diagnostics_residual_qq.png",
  width = 12,
  height = 7
)
if (write_outputs) {
  ggsave(
    filename = file.path(output_dir, "diagnostics_residual_qq.pdf"),
    plot = p_diagnostics_qq,
    width = 12,
    height = 7,
    units = "in",
    bg = "white"
  )
}
```

## 12. Starter Plots

### 12.1 Labelled Incorporation Time Course

```{r plot-labelled-timecourse, fig.width=14, fig.height=18}
labelled_curve_fdr_labels <- flux_analysis_summary |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  curve_label_positions(c("experiment", "feature_name")) |>
  left_join(
    curve_fdr_labels |>
      filter(readout == "labelled_incorporation") |>
      select(experiment, feature_name, curve_fdr_label),
    by = c("experiment", "feature_name")
  )

p_labelled_timecourse <- flux_analysis_summary |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  add_timepoint_plot_position() |>
  ggplot(aes(x = time_plot, y = mean_value, color = infection, group = infection)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.8) +
  geom_errorbar(
    aes(ymin = mean_value - se_value, ymax = mean_value + se_value),
    width = 0.15,
    linewidth = 0.4
  ) +
  geom_label(
    data = labelled_curve_fdr_labels,
    aes(x = label_x, y = label_y, label = curve_fdr_label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 2.4,
    linewidth = 0.15,
    label.r = grid::unit(0.06, "lines"),
    fill = "white",
    alpha = 0.85,
    color = "grey20",
    na.rm = TRUE
  ) +
  facet_wrap(vars(experiment, feature_name), scales = "free_y", ncol = 6) +
  scale_timepoint_x() +
  scale_color_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  labs(
    x = "Time after tracer addition (h)",
    y = "Labelled incorporation (pmol / sample)",
    color = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )

p_labelled_timecourse
save_plot_if_requested(
  p_labelled_timecourse,
  "labelled_incorporation_timecourse.png",
  width = 14,
  height = 18
)
```

### 12.2 AUC Ratio Summary

```{r plot-auc-ratio, fig.width=14, fig.height=6}
p_auc_ratio <- auc_ratio_summary |>
  filter(is.finite(infected_over_uninfected_auc)) |>
  ggplot(aes(x = feature_name, y = infected_over_uninfected_auc, fill = feature_type)) +
  geom_hline(yintercept = 1, linewidth = 0.4, color = "grey45") +
  geom_col(color = "black", width = 0.65) +
  facet_wrap(~ experiment, nrow = 1, scales = "free_x") +
  scale_fill_manual(
    values = c("LCB molecule" = "#E69F00", "Lipid species" = "#56B4E9")
  ) +
  labs(
    x = NULL,
    y = "AUC ratio: infected / uninfected",
    fill = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_blank()
  )

p_auc_ratio
save_plot_if_requested(
  p_auc_ratio,
  "auc_ratio_infected_over_uninfected.png",
  width = 14,
  height = 6
)
```

### 12.3 Labelled LCB Molecules

This plot is deliberately molecule-level. LCBs are not summed into a broad
class because `dhSph`, `Sph`, and `S1P` represent distinct sphingoid-base
molecules rather than acyl-chain variants of one lipid class.

```{r plot-lcb-labelled-species, fig.width=12, fig.height=6}
lcb_labelled_curve_fdr_labels <- flux_lcb_summary |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  mutate(species = as.character(species)) |>
  curve_label_positions(c("experiment", "species")) |>
  left_join(
    curve_fdr_labels |>
      filter(
        readout == "labelled_incorporation",
        feature_type == "LCB molecule"
      ) |>
      transmute(
        experiment,
        species = as.character(feature_name),
        curve_fdr_label
      ),
    by = c("experiment", "species")
  ) |>
  mutate(species = factor(species, levels = lcb_species_levels))

p_lcb_labelled_species <- flux_lcb_summary |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  mutate(species = droplevels(species)) |>
  add_timepoint_plot_position() |>
  ggplot(aes(x = time_plot, y = mean_value, color = infection, group = infection)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.8) +
  geom_errorbar(
    aes(ymin = mean_value - se_value, ymax = mean_value + se_value),
    width = 0.15,
    linewidth = 0.4
  ) +
  geom_label(
    data = lcb_labelled_curve_fdr_labels,
    aes(x = label_x, y = label_y, label = curve_fdr_label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 2.4,
    linewidth = 0.15,
    label.r = grid::unit(0.06, "lines"),
    fill = "white",
    alpha = 0.85,
    color = "grey20",
    na.rm = TRUE
  ) +
  facet_wrap(vars(experiment, species), scales = "free_y", ncol = 5) +
  scale_timepoint_x() +
  scale_color_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  labs(
    x = "Time after tracer addition (h)",
    y = "Labelled LCB molecule (pmol / sample)",
    color = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )

p_lcb_labelled_species
save_plot_if_requested(
  p_lcb_labelled_species,
  "lcb_labelled_species_timecourse.png",
  width = 12,
  height = 6
)
```

### 12.4 Unlabelled LCB Molecules

```{r plot-lcb-unlabelled-species, fig.width=12, fig.height=6}
lcb_unlabelled_curve_fdr_labels <- flux_lcb_summary |>
  filter(
    readout == "unlabelled_abundance",
    is_tracer_treated
  ) |>
  mutate(species = as.character(species)) |>
  curve_label_positions(c("experiment", "species")) |>
  left_join(
    curve_fdr_labels |>
      filter(
        readout == "unlabelled_abundance",
        feature_type == "LCB molecule"
      ) |>
      transmute(
        experiment,
        species = as.character(feature_name),
        curve_fdr_label
      ),
    by = c("experiment", "species")
  ) |>
  mutate(species = factor(species, levels = lcb_species_levels))

p_lcb_unlabelled_species <- flux_lcb_summary |>
  filter(
    readout == "unlabelled_abundance",
    is_tracer_treated
  ) |>
  mutate(species = droplevels(species)) |>
  add_timepoint_plot_position() |>
  ggplot(aes(x = time_plot, y = mean_value, color = infection, group = infection)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.8) +
  geom_errorbar(
    aes(ymin = mean_value - se_value, ymax = mean_value + se_value),
    width = 0.15,
    linewidth = 0.4
  ) +
  geom_label(
    data = lcb_unlabelled_curve_fdr_labels,
    aes(x = label_x, y = label_y, label = curve_fdr_label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 2.4,
    linewidth = 0.15,
    label.r = grid::unit(0.06, "lines"),
    fill = "white",
    alpha = 0.85,
    color = "grey20",
    na.rm = TRUE
  ) +
  facet_wrap(vars(experiment, species), scales = "free_y", ncol = 5) +
  scale_timepoint_x() +
  scale_color_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  labs(
    x = "Time after tracer addition (h)",
    y = "Unlabelled LCB molecule (pmol / sample)",
    color = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )

p_lcb_unlabelled_species
save_plot_if_requested(
  p_lcb_unlabelled_species,
  "lcb_unlabelled_species_timecourse.png",
  width = 12,
  height = 6
)
```

### 12.5 Unlabelled Abundance Time Course

```{r plot-unlabelled-timecourse, fig.width=14, fig.height=24}
unlabelled_curve_fdr_labels <- flux_analysis_summary |>
  filter(
    readout == "unlabelled_abundance",
    is_tracer_treated
  ) |>
  curve_label_positions(c("experiment", "feature_name")) |>
  left_join(
    curve_fdr_labels |>
      filter(readout == "unlabelled_abundance") |>
      select(experiment, feature_name, curve_fdr_label),
    by = c("experiment", "feature_name")
  )

p_unlabelled_timecourse <- flux_analysis_summary |>
  filter(
    readout == "unlabelled_abundance",
    is_tracer_treated
  ) |>
  add_timepoint_plot_position() |>
  ggplot(aes(x = time_plot, y = mean_value, color = infection, group = infection)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.8) +
  geom_errorbar(
    aes(ymin = mean_value - se_value, ymax = mean_value + se_value),
    width = 0.15,
    linewidth = 0.4
  ) +
  geom_label(
    data = unlabelled_curve_fdr_labels,
    aes(x = label_x, y = label_y, label = curve_fdr_label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 2.4,
    linewidth = 0.15,
    label.r = grid::unit(0.06, "lines"),
    fill = "white",
    alpha = 0.85,
    color = "grey20",
    na.rm = TRUE
  ) +
  facet_wrap(vars(experiment, feature_name), scales = "free_y", ncol = 6) +
  scale_timepoint_x() +
  scale_color_manual(values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")) +
  labs(
    x = "Time after tracer addition (h)",
    y = "Unlabelled abundance (pmol / sample)",
    color = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )

p_unlabelled_timecourse
save_plot_if_requested(
  p_unlabelled_timecourse,
  "unlabelled_abundance_timecourse.png",
  width = 14,
  height = 24
)
```

### 12.6 Export Individual Panels

Set `export_individual_panels: true` in the YAML parameters, or run with
`EXPORT_INDIVIDUAL_PANELS=true`, to save each faceted lipid panel as a separate
file. The default format is PNG. To write more than one format, set
`individual_panel_formats: "png,pdf"` or run with
`INDIVIDUAL_PANEL_FORMATS=png,pdf`.

```{r export-individual-panels}
plot_single_timecourse_panel <- function(
  panel_data,
  title,
  x_label,
  y_label,
  curve_fdr_label = NULL
) {
  panel_data <- add_timepoint_plot_position(panel_data)

  p <- ggplot(
    panel_data,
    aes(x = time_plot, y = mean_value, color = infection, group = infection)
  ) +
    geom_line(linewidth = 0.5) +
    geom_point(size = 1.8) +
    geom_errorbar(
      aes(
        ymin = pmax(mean_value - se_value, 0),
        ymax = mean_value + se_value
      ),
      width = 0.15,
      linewidth = 0.4
    ) +
    scale_timepoint_x() +
    scale_color_manual(
      values = c("Uninfected" = "#0072B2", "Infected" = "#D55E00")
    ) +
    labs(
      title = title,
      x = x_label,
      y = y_label,
      color = NULL
    ) +
    theme_classic(base_size = 10) +
    theme(
      plot.title = element_text(size = 11, face = "bold"),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      legend.position = "top"
    )

  if (!is.null(curve_fdr_label) && length(curve_fdr_label) > 0) {
    y_min <- min(pmax(panel_data$mean_value - panel_data$se_value, 0), na.rm = TRUE)
    y_max <- max(panel_data$mean_value + panel_data$se_value, na.rm = TRUE)
    y_range <- ifelse(
      is.finite(y_max - y_min) && y_max > y_min,
      y_max - y_min,
      pmax(abs(y_max), 1)
    )
    label_y <- y_max + 0.08 * y_range
    label_time <- min(panel_data$time_h, na.rm = TRUE)
    label_x <- timepoint_plot_breaks[match(label_time, timepoint_levels)]

    if (is.finite(label_y) && is.finite(label_x)) {
      p <- p +
        annotate(
          "label",
          x = label_x,
          y = label_y,
          label = curve_fdr_label[1],
          hjust = 0,
          vjust = 1,
          size = 2.7,
          linewidth = 0.15,
          label.r = grid::unit(0.06, "lines"),
          fill = "white",
          alpha = 0.85,
          color = "grey20"
        )
    }
  }

  p
}

individual_panel_exports <- tibble(
  panel_type = character(),
  experiment = character(),
  panel = character(),
  format = character(),
  file = character()
)

record_panel_exports <- function(panel_type, experiment, panel, path_stem) {
  bind_rows(
    individual_panel_exports,
    tibble(
      panel_type = panel_type,
      experiment = as.character(experiment),
      panel = as.character(panel),
      format = individual_panel_formats,
      file = file.path(output_dir, paste0(path_stem, ".", individual_panel_formats))
    )
  )
}

labelled_panel_index <- flux_analysis_summary |>
  filter(readout == "labelled_incorporation", is_tracer_treated) |>
  distinct(experiment, feature_type, feature_name) |>
  arrange(experiment, feature_name)

for (i in seq_len(nrow(labelled_panel_index))) {
  panel_experiment <- labelled_panel_index$experiment[i]
  panel_feature_type <- labelled_panel_index$feature_type[i]
  panel_feature <- labelled_panel_index$feature_name[i]

  panel_data <- flux_analysis_summary |>
    filter(
      readout == "labelled_incorporation",
      is_tracer_treated,
      experiment == panel_experiment,
      feature_name == panel_feature
    )

  panel_curve_fdr_label <- curve_fdr_labels |>
    filter(
      readout == "labelled_incorporation",
      experiment == panel_experiment,
      feature_name == panel_feature
    ) |>
    pull(curve_fdr_label)

  path_stem <- file.path(
    "individual_panels",
    "labelled_incorporation",
    paste(
      safe_file_part(panel_experiment),
      safe_file_part(panel_feature),
      sep = "_"
    )
  )

  p_panel <- plot_single_timecourse_panel(
    panel_data = panel_data,
    title = paste(panel_experiment, panel_feature, "labelled incorporation"),
    x_label = "Time after tracer addition (h)",
    y_label = "Labelled incorporation (pmol / sample)",
    curve_fdr_label = panel_curve_fdr_label
  )

  save_individual_panel_if_requested(p_panel, path_stem, width = 4.2, height = 3.4)
  individual_panel_exports <- record_panel_exports(
    "labelled_incorporation",
    panel_experiment,
    paste(panel_feature_type, panel_feature, sep = ": "),
    path_stem
  )
}

unlabelled_panel_index <- flux_analysis_summary |>
  filter(readout == "unlabelled_abundance", is_tracer_treated) |>
  distinct(experiment, feature_type, feature_name) |>
  arrange(experiment, feature_name)

for (i in seq_len(nrow(unlabelled_panel_index))) {
  panel_experiment <- unlabelled_panel_index$experiment[i]
  panel_feature_type <- unlabelled_panel_index$feature_type[i]
  panel_feature <- unlabelled_panel_index$feature_name[i]

  panel_data <- flux_analysis_summary |>
    filter(
      readout == "unlabelled_abundance",
      is_tracer_treated,
      experiment == panel_experiment,
      feature_name == panel_feature
    )

  panel_curve_fdr_label <- curve_fdr_labels |>
    filter(
      readout == "unlabelled_abundance",
      experiment == panel_experiment,
      feature_name == panel_feature
    ) |>
    pull(curve_fdr_label)

  path_stem <- file.path(
    "individual_panels",
    "unlabelled_abundance",
    paste(
      safe_file_part(panel_experiment),
      safe_file_part(panel_feature),
      sep = "_"
    )
  )

  p_panel <- plot_single_timecourse_panel(
    panel_data = panel_data,
    title = paste(panel_experiment, panel_feature, "unlabelled abundance"),
    x_label = "Time after tracer addition (h)",
    y_label = "Unlabelled abundance (pmol / sample)",
    curve_fdr_label = panel_curve_fdr_label
  )

  save_individual_panel_if_requested(p_panel, path_stem, width = 4.2, height = 3.4)
  individual_panel_exports <- record_panel_exports(
    "unlabelled_abundance",
    panel_experiment,
    paste(panel_feature_type, panel_feature, sep = ": "),
    path_stem
  )
}

if (write_outputs && export_individual_panels) {
  write.csv(
    individual_panel_exports,
    file.path(output_dir, "individual_panel_exports.csv"),
    row.names = FALSE
  )
}

individual_panel_exports
```

### 12.7 Export Prism-Style Labelled Incorporation Tables

This export creates one Prism-style workbook for each current labelled
incorporation panel. The tables use absolute `pmol / sample` values. Column A
contains timepoints, columns B-D contain uninfected replicates (`-Sne`), and
columns E-G contain infected replicates (`+Sne`).

```{r export-prism-labelled-tables}
prism_labelled_dir <- file.path(
  output_dir,
  "prism_labelled_incorporation_tables"
)

write_prism_labelled_table <- function(panel_data, xlsx_path) {
  timepoints <- sort(unique(panel_data$time_h))
  replicate_levels <- sort(unique(as.character(panel_data$replicate)))

  if (length(replicate_levels) < 3) {
    replicate_levels <- union(c("1", "2", "3"), replicate_levels)
  }

  replicate_levels <- replicate_levels[seq_len(min(3, length(replicate_levels)))]

  make_group_matrix <- function(infection_label) {
    panel_data |>
      filter(infection == infection_label) |>
      mutate(
        replicate = factor(as.character(replicate), levels = replicate_levels),
        time_h = factor(time_h, levels = timepoints)
      ) |>
      select(time_h, replicate, value) |>
      tidyr::complete(time_h, replicate) |>
      arrange(time_h, replicate) |>
      tidyr::pivot_wider(names_from = replicate, values_from = value) |>
      select(dplyr::all_of(replicate_levels)) |>
      as.matrix()
  }

  wb <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(wb, "Sheet1")

  openxlsx::writeData(wb, "Sheet1", "-Sne", startCol = 2, startRow = 1, colNames = FALSE)
  openxlsx::writeData(wb, "Sheet1", "+Sne", startCol = 5, startRow = 1, colNames = FALSE)
  openxlsx::writeData(wb, "Sheet1", timepoints, startCol = 1, startRow = 2, colNames = FALSE)
  openxlsx::writeData(
    wb,
    "Sheet1",
    make_group_matrix("Uninfected"),
    startCol = 2,
    startRow = 2,
    colNames = FALSE
  )
  openxlsx::writeData(
    wb,
    "Sheet1",
    make_group_matrix("Infected"),
    startCol = 5,
    startRow = 2,
    colNames = FALSE
  )

  openxlsx::setColWidths(wb, "Sheet1", cols = 1:7, widths = 12)
  openxlsx::saveWorkbook(wb, xlsx_path, overwrite = TRUE)
  invisible(xlsx_path)
}

prism_labelled_panel_index <- flux_analysis_readouts |>
  filter(
    readout == "labelled_incorporation",
    is_tracer_treated
  ) |>
  distinct(experiment, tracer, feature_type, feature_name) |>
  arrange(experiment, feature_name)

if (write_outputs) {
  dir.create(prism_labelled_dir, recursive = TRUE, showWarnings = FALSE)
}

prism_labelled_table_exports <- tibble()

for (i in seq_len(nrow(prism_labelled_panel_index))) {
  panel_experiment <- prism_labelled_panel_index$experiment[i]
  panel_tracer <- prism_labelled_panel_index$tracer[i]
  panel_feature_type <- prism_labelled_panel_index$feature_type[i]
  panel_feature <- prism_labelled_panel_index$feature_name[i]
  file_stem <- paste(
    safe_file_part(panel_experiment),
    safe_file_part(panel_feature),
    sep = "_"
  )
  xlsx_path <- file.path(prism_labelled_dir, paste0(file_stem, ".xlsx"))

  panel_data <- flux_analysis_readouts |>
    filter(
      readout == "labelled_incorporation",
      is_tracer_treated,
      experiment == panel_experiment,
      feature_name == panel_feature
    )

  if (write_outputs) {
    write_prism_labelled_table(panel_data, xlsx_path)
  }

  prism_labelled_table_exports <- bind_rows(
    prism_labelled_table_exports,
    tibble(
      panel_type = "labelled_incorporation",
      experiment = as.character(panel_experiment),
      tracer = as.character(panel_tracer),
      feature_type = as.character(panel_feature_type),
      feature_name = as.character(panel_feature),
      unit = "pmol_per_sample",
      prism_group_left = "-Sne",
      prism_group_right = "+Sne",
      file = xlsx_path
    )
  )
}

if (write_outputs) {
  write.csv(
    prism_labelled_table_exports,
    file.path(prism_labelled_dir, "prism_labelled_table_exports.csv"),
    row.names = FALSE
  )
  openxlsx::write.xlsx(
    prism_labelled_table_exports,
    file.path(prism_labelled_dir, "prism_labelled_table_exports.xlsx"),
    overwrite = TRUE
  )
}

prism_labelled_table_exports
```

## 13. Statistical Markdown Exports

This chunk writes reviewer-friendly Markdown summaries of the statistical
methods and current results whenever `write_outputs` is enabled.

```{r statistical-markdown-exports}
format_result_number <- function(x, digits = 3) {
  vapply(x, function(value) {
    if (is.na(value)) {
      "NA"
    } else if (is.numeric(value) && is.finite(value) && value != 0 && abs(value) < 10^-digits) {
      formatC(value, format = "e", digits = 2)
    } else if (is.numeric(value) && is.finite(value)) {
      formatC(value, format = "f", digits = digits)
    } else {
      as.character(value)
    }
  }, character(1))
}

format_p_value <- function(x) {
  vapply(x, function(value) {
    if (is.na(value)) {
      "NA"
    } else if (value < 0.001) {
      formatC(value, format = "e", digits = 2)
    } else {
      formatC(value, format = "f", digits = 3)
    }
  }, character(1))
}

as_markdown_table <- function(data) {
  if (nrow(data) == 0) {
    return("_No rows matched this criterion._")
  }

  table_data <- data
  table_data[] <- lapply(table_data, as.character)

  c(
    paste("|", paste(names(table_data), collapse = " | "), "|"),
    paste("|", paste(rep("---", ncol(table_data)), collapse = " | "), "|"),
    apply(
      table_data,
      1,
      function(row) paste("|", paste(row, collapse = " | "), "|")
    )
  )
}

write_markdown_if_requested <- function(lines, filename) {
  if (write_outputs) {
    stats_dir <- file.path(output_dir, "statistical_results_markdown")
    dir.create(stats_dir, recursive = TRUE, showWarnings = FALSE)
    writeLines(lines, file.path(stats_dir, filename))
  }
  invisible(lines)
}

auc_table <- auc_ratio_summary |>
  transmute(
    Experiment = as.character(experiment),
    `Feature type` = as.character(feature_type),
    Feature = as.character(feature_name),
    `Infected / uninfected AUC` = format_result_number(
      infected_over_uninfected_auc
    )
  )

timepoint_significant <- timepoint_tests |>
  filter(p_fdr < 0.05) |>
  transmute(
    Experiment = as.character(experiment),
    `Feature type` = as.character(feature_type),
    Feature = as.character(feature_name),
    `Time (h)` = format_result_number(time_h, digits = 1),
    Contrast = contrast,
    `Infected / uninfected` = format_result_number(
      ratio_infected_over_uninfected
    ),
    `95% ratio CI` = paste0(
      format_result_number(ratio_lower_95),
      "-",
      format_result_number(ratio_upper_95)
    ),
    `t ratio` = format_result_number(t_ratio, digits = 2),
    FDR = format_p_value(p_fdr)
  )

timepoint_primary_counts <- timepoint_tests |>
  filter(p_fdr < 0.05) |>
  count(experiment, name = "n_primary_experiment_fdr")

timepoint_panel_counts <- timepoint_tests_panel_fdr |>
  filter(p_fdr < 0.05) |>
  count(experiment, name = "n_panel_fdr")

timepoint_count_table <- full_join(
  timepoint_primary_counts,
  timepoint_panel_counts,
  by = "experiment"
) |>
  mutate(
    n_primary_experiment_fdr = replace_na(n_primary_experiment_fdr, 0L),
    n_panel_fdr = replace_na(n_panel_fdr, 0L)
  ) |>
  transmute(
    Experiment = as.character(experiment),
    `Significant tests, experiment-level FDR` = n_primary_experiment_fdr,
    `Significant tests, panel-level FDR sensitivity` = n_panel_fdr
  )

infection_time_significant <- infection_time_models |>
  filter(p_fdr_infection_time < 0.05) |>
  transmute(
    Experiment = as.character(experiment),
    `Feature type` = as.character(feature_type),
    Feature = as.character(feature_name),
    `Raw p, infection:time` = format_p_value(p_infection_time),
    `FDR, infection:time` = format_p_value(p_fdr_infection_time)
  )

zero_nearzero_nonzero <- zero_nearzero_summary |>
  filter(near_zero_pct > 0) |>
  transmute(
    Experiment = as.character(experiment),
    `Feature type` = as.character(feature_type),
    Feature = as.character(feature_name),
    n = n,
    `Zero n` = zero_n,
    `Near-zero %` = format_result_number(near_zero_pct, digits = 1)
  )

overview_lines <- c(
  "# Isotope Tracer Statistical Results Overview",
  "",
  "This static summary describes the current Pal-d3 and d7-dhSph tracer analyses after the species-level correction.",
  "",
  "LCBs are treated as individual molecule-level features (`dhSph`, `Sph`, `S1P`, and labelled forms). Non-LCB sphingolipid readouts are treated as individual labelled or unlabelled molecular species rather than class totals.",
  "",
  "The statistical tests use:",
  "",
  "```text",
  paste0("y = log10(value + ", pseudovalue, ")"),
  "```",
  "",
  "Plots remain on untransformed absolute `pmol / sample` values for biological readability. Class-level totals are retained only as supporting/QC outputs.",
  "",
  "Timepoint pairwise tests use `emmeans` for the explicit contrast `Infected - Uninfected`. The primary FDR correction is Benjamini-Hochberg within each tracer experiment across all tested labelled species and sampled timepoints.",
  "",
  "Species with zero or near-zero labelled signal can produce perfect-fit model warnings; inspect `diagnostics_zero_nearzero_summary.csv` and residual diagnostic plots before over-interpreting those panels.",
  "",
  "## AUC Summary",
  "",
  "```text",
  "AUC = sum((t[i+1] - t[i]) * (y[i] + y[i+1]) / 2)",
  "AUC ratio = mean AUC infected / mean AUC uninfected",
  "```",
  "",
  as_markdown_table(auc_table),
  "",
  "## Timepoint FDR Counts",
  "",
  as_markdown_table(timepoint_count_table),
  "",
  "## Significant Timepoint Pairwise Comparisons",
  "",
  as_markdown_table(timepoint_significant),
  "",
  "## Significant Infection-by-Time Interactions",
  "",
  as_markdown_table(infection_time_significant),
  "",
  "## Zero And Near-Zero Labelled Features",
  "",
  as_markdown_table(zero_nearzero_nonzero)
)

auc_lines <- c(
  "# AUC Results",
  "",
  "AUC is calculated on untransformed absolute `pmol / sample` values using the trapezoid rule across the observed 0.5-8 h time course.",
  "",
  "The reported ratio is mean infected AUC divided by mean uninfected AUC for each labelled species-level panel.",
  "",
  as_markdown_table(auc_table)
)

timepoint_lines <- c(
  "# Timepoint Pairwise Comparison Results",
  "",
  "At each timepoint and species-level feature, infected and uninfected tracer-treated samples are compared using:",
  "",
  "```text",
  paste0("model: log10(value + ", pseudovalue, ") ~ infection"),
  "pairwise contrast: Infected - Uninfected",
  "infected / uninfected = 10^(emmeans contrast estimate)",
  "```",
  "",
  "The primary p-value shown below is Benjamini-Hochberg FDR corrected within each tracer experiment across all tested labelled species and sampled timepoints.",
  "",
  "A secondary panel-level FDR sensitivity table is also written to `timepoint_pairwise_emmeans_panel_fdr_labelled_incorporation.csv`.",
  "",
  as_markdown_table(timepoint_significant)
)

infection_time_lines <- c(
  "# Infection-by-Time Model Results",
  "",
  "For each labelled and unlabelled species-level panel, the curve-level model is:",
  "",
  "```text",
  paste0("log10(value + ", pseudovalue, ") ~ infection * time_f"),
  "```",
  "",
  "The line-plot annotation uses the BH-adjusted `infection:time_f` p-value, reported as `curve FDR`. The table below lists labelled incorporation panels with FDR < 0.05 for the infection-by-time interaction.",
  "",
  as_markdown_table(infection_time_significant)
)

write_markdown_if_requested(overview_lines, "00_statistical_results_overview.md")
write_markdown_if_requested(auc_lines, "01_auc_results.md")
write_markdown_if_requested(timepoint_lines, "02_timepoint_tests.md")
write_markdown_if_requested(infection_time_lines, "03_infection_time_models.md")

list(
  auc_rows = nrow(auc_table),
  significant_timepoint_rows = nrow(timepoint_significant),
  significant_infection_time_rows = nrow(infection_time_significant),
  near_zero_feature_rows = nrow(zero_nearzero_nonzero)
)
```

## 14. Working Interpretation Notes

```{r interpretation-notes}
analysis_notes <- data.frame(
  point = c(
    "Primary unit",
    "Labelled and unlabelled readouts",
    "LCB handling",
    "Curve FDR annotation",
    "Pal-d3 interpretation",
    "d7-dhSph interpretation",
    "AUC use",
    "Timepoint use",
    "Class-level totals"
  ),
  note = c(
    "Use absolute pmol / sample as the primary analysis unit. The normalized block can be imported for sensitivity checks.",
    "Keep labelled incorporation and unlabelled abundance side by side at species level because they answer different biological questions.",
    "Do not sum LCBs into one broad class. Treat dhSph, Sph, S1P, and labelled LCBs as molecule-level features.",
    "The curve FDR shown on line plots is the BH-adjusted infection-by-time interaction p-value from log10(value + 0.01) ~ infection * time_f.",
    "Pal-d3 reports apparent entry from supplied palmitate into the de novo sphingolipid pathway.",
    "d7-dhSph reports handling of a supplied sphingoid base. Inspect d7-dhSph, d7-Sph, and d7-S1P separately.",
    "Use AUC as a compact species-level summary of labelled signal across the observed time course.",
    "Use experiment-level BH-FDR for timepoint tests across species/timepoint comparisons within each tracer experiment.",
    "Class-level totals are retained as supporting/QC outputs only; they are not the primary tracer analysis features."
  ),
  stringsAsFactors = FALSE
)

write_csv_if_requested(analysis_notes, "analysis_notes.csv")
analysis_notes
```

## 15. Session Info

```{r session-info}
sessionInfo()
```
